---
title: "Riley's Personal Analysis v2"
output: html_document
---

```{r}
wineData = read.csv("~/STAT394-Group-Project/extracted data.csv", header=TRUE)
summary(wineData)
```
```{r}
library(reshape2)
library(ggplot2)
wineData$Cultivar = c(rep("Mencia",20),rep("Alicante Bouschet",10))
wineMelt = melt(wineData[,-c(1,2)],
                id.vars = "Cultivar",
                variable.name = "Measure")
ggplot(data=wineMelt, aes(x=Measure, y=value)) +
  geom_boxplot(aes(col=Measure), notch = FALSE) +
  scale_y_log10() +
  facet_grid(~Cultivar) +
  ylab("Values")
```

```{r}
apply(wineData[1:20,3:25],2,shapiro.test)
```

First try PCA on whole dataset, look for cultivar separation.

```{r}
wineDataVariables = wineData[,3:25]


wineDataPCA = prcomp(wineDataVariables, center = TRUE, scale = TRUE)$x

summary(prcomp(wineDataVariables, center = TRUE, scale = TRUE))
# first 5 PCs have stdev>1, drop rest
wineDataPCA = wineDataPCA[,1:5]
summary(wineDataPCA)
wineDataPCADF = as.data.frame(wineDataPCA)
apply(wineDataPCADF[1:20,],2,shapiro.test)
```
```{r}
# mahalanobis to find surprising obs from model assuming no groups
PCAMahalanobis = mahalanobis(wineDataPCADF,center = colMeans(wineDataPCADF),cov = cov(wineDataPCADF))
PCAMahalanobis
surprise = pchisq(PCAMahalanobis,df=5,lower.tail = FALSE)
surprise
surpriseFull = ifelse(surprise <= 0.01, "Very Surprising",
                      ifelse(surprise <= 0.05, "Surprising",
                             ifelse(surprise <= 0.1, "Slightly Surprising", round(surprise,4))))
surpriseFull
```
```{r}
# check if AB obs are consistent with M distribution
PCAFullMenciaMahal = mahalanobis(wineDataPCADF, center=colMeans(wineDataPCADF[1:20,]), cov=cov(wineDataPCADF[1:20,]))
PCAFullMenciaMahal
surpriseMencia = pchisq(PCAFullMenciaMahal, df=5, lower.tail=FALSE)
surpriseMenciaWords = ifelse(surpriseMencia <= 0.01, "Very Surprising",
                      ifelse(surpriseMencia <= 0.05, "Surprising",
                             ifelse(surpriseMencia <= 0.1, "Slightly Surprising", round(surpriseMencia,4))))
surpriseMenciaWords
surpriseMencia
```
```{r}
# check if M obs are consistent with AB distribution
PCAFullAlicanteMahal = mahalanobis(wineDataPCADF, center=colMeans(wineDataPCADF[21:30,]), cov=cov(wineDataPCADF[21:30,]))
PCAFullAlicanteMahal
surpriseAlicante = pchisq(PCAFullAlicanteMahal, df=5, lower.tail=FALSE)
surpriseAlicanteWords = ifelse(surpriseAlicante <= 0.01, "Very Surprising",
                      ifelse(surpriseAlicante <= 0.05, "Surprising",
                             ifelse(surpriseAlicante <= 0.1, "Slightly Surprising", round(surpriseAlicante,4))))
surpriseAlicanteWords
surpriseAlicante
```

```{r}
# PCs from  full dataset, mahalanobis from individual cultivar distributions excluding the individual observation

surpriseCulExc = rep(NA,30)
mahalCulExc = rep(NA,30)
for(i in 1:20){
  wineTemp = wineDataPCADF[1:20,][-i,]
  muTemp = colMeans(wineTemp)
  covTemp = cov(wineTemp)
  mahalCulExc[i] = mahalanobis(wineDataPCADF[i,], center = muTemp, cov=covTemp)
  surpriseCulExc[i] = pchisq(mahalCulExc[i], df=5, lower.tail=FALSE)
}
for(i in 21:30){
  iAdj = i - 20
  wineTemp = wineDataPCADF[21:30,][-iAdj,]
  muTemp = colMeans(wineTemp)
  covTemp = cov(wineTemp)
  mahalCulExc[i] = mahalanobis(wineDataPCADF[i,], center = muTemp, cov=covTemp)
  surpriseCulExc[i] = pchisq(mahalCulExc[i], df=5, lower.tail=FALSE)
}
mahalCulExc
surpriseCulExcWords =  ifelse(surpriseCulExc <= 0.01, "Very Surprising",
                      ifelse(surpriseCulExc <= 0.05, "Surprising",
                             ifelse(surpriseCulExc <= 0.1, "Slightly Surprising", round(surpriseCulExc,4))))
surpriseCulExcWords
round(surpriseCulExc,6)
```

```{r}
#  playing around with plots
library(ggplot2)
library(ggcorrplot)
library(GGally)
plot(wineDataPCADF$PC1,
     wineDataPCADF$PC2,
     col = c(rep("red",20),rep("blue",10)),
     pch = c(LETTERS,letters[1:4]))
```

```{r}
# playing with graphs for within AB subgroup
plot(wineDataPCADF[21:30,]$PC1,
     wineDataPCADF[21:30,]$PC2,
     col = c("green",rep("black",5),rep("green",4)),
     pch = as.character(c(1:9,0)))
```

```{r}
# mahalanobis analysis of subclusters within AB (without exclusion) (all PCs) (PCs from full dataset)
wineAB1 = wineDataPCADF[c(21,27:30),]
wineAB2 = wineDataPCADF[22:26,]
muAB1 = colMeans(wineAB1)
muAB2 = colMeans(wineAB2)
sigAB1 = cov(wineAB1)
sigAB2 = cov(wineAB2)
# mahalanobis(wineAB1,center = muAB1, cov = sigAB1)
# samples too small to calculate mahalanobis distance on 5 variables. PC5 shows no clustering, drop and reattempt
```


```{r}
# mahalanobis analysis of subclusters within AB (without exclusion) (PC1-4) (PCs from full dataset)
wineAB1 = wineDataPCADF[c(21,27:30),1:4]
wineAB2 = wineDataPCADF[22:26,1:4]
muAB1 = colMeans(wineAB1)
muAB2 = colMeans(wineAB2)
sigAB1 = cov(wineAB1)
sigAB2 = cov(wineAB2)
mahalanobis(wineAB1,center = muAB1, cov = sigAB1)
# still too many variables for number of obs, gives constant dM. repeat with just PC1,PC2
```


```{r}
# mahalanobis analysis of subclusters within AB (without exclusion) (PC1-3) (PCs from full dataset)
wineAB1 = wineDataPCADF[c(21,27:30),1:3]
wineAB2 = wineDataPCADF[22:26,1:3]
muAB1 = colMeans(wineAB1)
muAB2 = colMeans(wineAB2)
sigAB1 = cov(wineAB1)
sigAB2 = cov(wineAB2)
mahalanobis(wineAB1,center = muAB1, cov = sigAB1)
mahalanobis(wineAB2,center = muAB1, cov = sigAB1)
mahalanobis(wineAB2,center = muAB2, cov = sigAB2)
mahalanobis(wineAB1,center = muAB2, cov = sigAB2)

pchisq(mahalanobis(wineAB1,center = muAB1, cov = sigAB1), df=2, lower.tail = FALSE)
pchisq(mahalanobis(wineAB2,center = muAB1, cov = sigAB1), df=2, lower.tail = FALSE)
pchisq(mahalanobis(wineAB2,center = muAB2, cov = sigAB2), df=2, lower.tail = FALSE)
pchisq(mahalanobis(wineAB1,center = muAB2, cov = sigAB2), df=2, lower.tail = FALSE)

```

```{r}
# mahalanobis analysis of subclusters within AB (without exclusion) (PC1-2) (PCs from full dataset)
wineAB1 = wineDataPCADF[c(21,27:30),1:2]
wineAB2 = wineDataPCADF[22:26,1:2]
muAB1 = colMeans(wineAB1)
muAB2 = colMeans(wineAB2)
sigAB1 = cov(wineAB1)
sigAB2 = cov(wineAB2)
mahalanobis(wineAB1,center = muAB1, cov = sigAB1)
mahalanobis(wineAB2,center = muAB1, cov = sigAB1)
mahalanobis(wineAB2,center = muAB2, cov = sigAB2)
mahalanobis(wineAB1,center = muAB2, cov = sigAB2)


pchisq(mahalanobis(wineAB1,center = muAB1, cov = sigAB1), df=2, lower.tail = FALSE)
pchisq(mahalanobis(wineAB2,center = muAB1, cov = sigAB1), df=2, lower.tail = FALSE)
pchisq(mahalanobis(wineAB2,center = muAB2, cov = sigAB2), df=2, lower.tail = FALSE)
pchisq(mahalanobis(wineAB1,center = muAB2, cov = sigAB2), df=2, lower.tail = FALSE)

```
```{r}
# factor loadings
prcomp(wineDataVariables, center = TRUE, scale = TRUE)$rotation[,1:2]
library(ggbiplot)
library(rlang)
ggbiplot(prcomp(wineDataVariables, center = TRUE, scale = TRUE), obs.scale = 1, var.scale = 1, alpha = 0.1)
```

```{r}
# mahalanobis analysis of subclusters within AB (with exclusion) (PC1-2) (PCs from full dataset)
wineAB1 = wineDataPCADF[c(21,27:30),1:2]
wineAB2 = wineDataPCADF[22:26,1:2]
wineAB = wineDataPCADF[21:30,1:2]

surpFromAB1Exc = rep(NA,10)
surpFromAB2Exc = rep(NA,10)
ind = 0
for(i in c(1,7:10)){
  ind = ind + 1
  wineAB1Temp = wineAB1[-ind,]
  iTemp = i + 20
  mahalAB1Temp = mahalanobis(wineAB1[ind,], center = colMeans(wineAB1Temp), cov = cov(wineAB1Temp))
  mahalAB2Temp = mahalanobis(wineAB1[ind,], center = colMeans(wineAB2), cov = cov(wineAB2))
  surpFromAB1Exc[i] = pchisq(mahalAB1Temp, df=2, lower.tail=FALSE)
  surpFromAB2Exc[i] = pchisq(mahalAB2Temp, df=2, lower.tail = FALSE)
}

ind = 0
for(i in 2:6){
  ind = ind + 1
  wineAB2Temp = wineAB2[-ind,]
  iTemp = i + 20
  mahalAB1Temp = mahalanobis(wineAB2[ind,], center = colMeans(wineAB1), cov = cov(wineAB1))
  mahalAB2Temp = mahalanobis(wineAB2[ind,], center = colMeans(wineAB2Temp), cov = cov(wineAB2Temp))
  surpFromAB1Exc[i] = pchisq(mahalAB1Temp, df=2, lower.tail=FALSE)
  surpFromAB2Exc[i] = pchisq(mahalAB2Temp, df=2, lower.tail = FALSE)
}
surpFromAB1Exc
surpFromAB2Exc

surpFromAB1ExcWords =  ifelse(surpFromAB1Exc <= 0.01, "Very Surprising",
                      ifelse(surpFromAB1Exc <= 0.05, "Surprising",
                             ifelse(surpFromAB1Exc <= 0.1, "Slightly Surprising", round(surpFromAB1Exc,4))))

surpFromAB2ExcWords =  ifelse(surpFromAB2Exc <= 0.01, "Very Surprising",
                      ifelse(surpFromAB2Exc <= 0.05, "Surprising",
                             ifelse(surpFromAB2Exc <= 0.1, "Slightly Surprising", round(surpFromAB2Exc,4))))

surpFromAB1ExcWords
surpFromAB2ExcWords
```

```{r}
# mahalanobis analysis of subclusters within AB (with exclusion) (PC1-2) (PCs from full dataset) (switch 21 into cluster 2)
wineAB1 = wineDataPCADF[27:30,1:2]
wineAB2 = wineDataPCADF[21:26,1:2]
wineAB = wineDataPCADF[21:30,1:2]

surpFromAB1Exc = rep(NA,10)
surpFromAB2Exc = rep(NA,10)
ind = 0
for(i in 7:10){
  ind = ind + 1
  wineAB1Temp = wineAB1[-ind,]
  iTemp = i + 20
  mahalAB1Temp = mahalanobis(wineAB1[ind,], center = colMeans(wineAB1Temp), cov = cov(wineAB1Temp))
  mahalAB2Temp = mahalanobis(wineAB1[ind,], center = colMeans(wineAB2), cov = cov(wineAB2))
  surpFromAB1Exc[i] = pchisq(mahalAB1Temp, df=2, lower.tail=FALSE)
  surpFromAB2Exc[i] = pchisq(mahalAB2Temp, df=2, lower.tail = FALSE)
}

ind = 0
for(i in 1:6){
  ind = ind + 1
  wineAB2Temp = wineAB2[-ind,]
  iTemp = i + 20
  mahalAB1Temp = mahalanobis(wineAB2[ind,], center = colMeans(wineAB1), cov = cov(wineAB1))
  mahalAB2Temp = mahalanobis(wineAB2[ind,], center = colMeans(wineAB2Temp), cov = cov(wineAB2Temp))
  surpFromAB1Exc[i] = pchisq(mahalAB1Temp, df=2, lower.tail=FALSE)
  surpFromAB2Exc[i] = pchisq(mahalAB2Temp, df=2, lower.tail = FALSE)
}
surpFromAB1Exc
surpFromAB2Exc

surpFromAB1ExcWords =  ifelse(surpFromAB1Exc <= 0.01, "Very Surprising",
                      ifelse(surpFromAB1Exc <= 0.05, "Surprising",
                             ifelse(surpFromAB1Exc <= 0.1, "Slightly Surprising", round(surpFromAB1Exc,4))))

surpFromAB2ExcWords =  ifelse(surpFromAB2Exc <= 0.01, "Very Surprising",
                      ifelse(surpFromAB2Exc <= 0.05, "Surprising",
                             ifelse(surpFromAB2Exc <= 0.1, "Slightly Surprising", round(surpFromAB2Exc,4))))

surpFromAB1ExcWords
surpFromAB2ExcWords
```

Results write-up sections

```{r}
# PCA
PCs = prcomp(wineDataVariables, center = TRUE, scale = TRUE)
screeplot(PCs,type = "lines") +
  abline(h=1, col="red")
```

```{r}
# Shapiro-Wilk test of PC1-5 aggregate and by cultivar
aggShap = rep(NA,5)
menShap = rep(NA,5)
abShap = rep(NA,5)
for(i in 1:5){
  aggShap[i] = shapiro.test(wineDataPCADF[,i])$p.value
  menShap[i] = shapiro.test(wineDataPCADF[1:20,i])$p.value
  abShap[i] = shapiro.test(wineDataPCADF[21:30,i])$p.value
}
shapDF = data.frame(Component = c("PC1","PC2","PC3","PC4","PC5"), Aggregate = aggShap, Mencia = menShap, AlicanteBouschet = abShap)
colnames(shapDF)[4] = "Alicante Bouschet"

for(i in 1:5){
  for(j in 2:4){
    charTemp = as.character(round(as.numeric(shapDF[i,j]),4))
    shapDF[i,j] = ifelse(shapDF[i,j] <= 0.01, paste0(charTemp," ***"),
                        ifelse(shapDF[i,j] <= 0.05, paste0(charTemp, " **"),
                               ifelse(shapDF[i,j] <=0.1, paste0(charTemp, " *"),
                                      charTemp)))
  }
}
pander(shapDF)
```


```{r}
# graphical of PCs, full dataset
plot(wineDataPCADF$PC1,
     wineDataPCADF$PC2,
     col = c(rep("red",20),rep("blue",10)),
     lwd = 3) + text(wineDataPCADF$PC1+0.1, wineDataPCADF$PC2-0.2, wineData$Sample, cex = 0.5)
# ggplot(wineDataPCADF, aes(x=PC1,y=PC2)) +
#   geom_point(colour = c(rep("red",20),rep("blue",10))) +
#   geom_text(aes(x=PC1,y=PC2, label = wineData$Sample, cex=0.5))
```


```{r}
# surprising results for mahalanobis: full model, cultivar split, cultivar split with exclusion
library(pander)
presFullModel = rep(NA,30)
presMenInc = rep(NA,30)
presABInc = rep(NA,30)
presSplitExc = rep(NA,30)
for(i in 1:30){
  if(surprise[i] < 0.0001){
    presFullModel[i] = "< 0.0001 ***"
  } else if(surprise[i] <= 0.01){
    presFullModel[i] = paste0(round(surprise[i],4)," ***")
  } else if(surprise[i] <= 0.05){
    presFullModel[i] = paste0(round(surprise[i],4)," **")
  } else if(surprise[i] <= 0.1){
    presFullModel[i] = paste0(round(surprise[i],4)," *")
  } else {
    presFullModel[i] = as.character(round(surprise[i],4))
  }
  if(surpriseMencia[i] < 0.0001){
    presMenInc[i] = "< 0.0001 ***"
  } else if(surpriseMencia[i] <= 0.01){
    presMenInc[i] = paste0(round(surpriseMencia[i],4)," ***")
  } else if(surpriseMencia[i] <= 0.05){
    presMenInc[i] = paste0(round(surpriseMencia[i],4)," **")
  } else if(surpriseMencia[i] <= 0.1){
    presMenInc[i] = paste0(round(surpriseMencia[i],4)," *")
  } else {
    presMenInc[i] = as.character(round(surpriseMencia[i],4))
  }
  if(surpriseAlicante[i] < 0.0001){
    presABInc[i] = "< 0.0001 ***"
  } else if(surpriseAlicante[i] <= 0.01){
    presABInc[i] = paste0(round(surpriseAlicante[i],4)," ***")
  } else if(surpriseAlicante[i] <= 0.05){
    presABInc[i] = paste0(round(surpriseAlicante[i],4)," **")
  } else if(surpriseAlicante[i] <= 0.1){
    presABInc[i] = paste0(round(surpriseAlicante[i],4)," *")
  } else {
    presABInc[i] = as.character(round(surpriseAlicante[i],4))
  }
  if(surpriseCulExc[i] < 0.0001){
    presSplitExc[i] = "< 0.0001 ***"
  } else if(surpriseCulExc[i] <= 0.01){
    presSplitExc[i] = paste0(round(surpriseCulExc[i],4)," ***")
  } else if(surpriseCulExc[i] <= 0.05){
    presSplitExc[i] = paste0(round(surpriseCulExc[i],4)," **")
  } else if(surpriseCulExc[i] <= 0.1){
    presSplitExc[i] = paste0(round(surpriseCulExc[i],4)," *")
  } else {
    presSplitExc[i] = as.character(round(surpriseCulExc[i],4))
  }
}
mahalTable = data.frame(sample = wineData$Sample, fullModel = presFullModel, menciaInc = presMenInc, abInc = presABInc, splitExc = presSplitExc)
mahalTable$splitExc = gsub("2e-04","0.0002",mahalTable$splitExc)
colnames(mahalTable) = c("Sample","Full Dataset","Mencia","Alicante Bouschet","Cultivar Excluding Sample")
pander(mahalTable)
```

```{r}
# graph PC1-2 AB split
plot(wineDataPCADF[21:30,]$PC1,
     wineDataPCADF[21:30,]$PC2,
     col = c("green",rep("black",5),rep("green",4)),
     pch = as.character(c(1:9,0)))
```

```{r}
# mahalanobis AB PC1-3 without exclusion
library(pander)
presCluster1 = rep(NA,10)
presCluster2 = rep(NA,10)

wineAB1 = wineDataPCADF[c(21,27:30),1:3]
wineAB2 = wineDataPCADF[22:26,1:3]
muAB1 = colMeans(wineAB1)
muAB2 = colMeans(wineAB2)
sigAB1 = cov(wineAB1)
sigAB2 = cov(wineAB2)

clus1from1surp = pchisq(mahalanobis(wineAB1,center = muAB1, cov = sigAB1), df=2, lower.tail = FALSE)
clus2from1surp = pchisq(mahalanobis(wineAB2,center = muAB1, cov = sigAB1), df=2, lower.tail = FALSE)
clus2from2surp = pchisq(mahalanobis(wineAB2,center = muAB2, cov = sigAB2), df=2, lower.tail = FALSE)
clus1from2surp = pchisq(mahalanobis(wineAB1,center = muAB2, cov = sigAB2), df=2, lower.tail = FALSE)

clus1Surp = c(clus1from1surp[1],clus2from1surp,clus1from1surp[2:5])
clus2Surp = c(clus1from2surp[1],clus2from2surp,clus1from2surp[2:5])



for(i in 1:10){
  if(clus1Surp[i] < 0.0001){
    presCluster1[i] = "< 0.0001 ***"
  } else if(clus1Surp[i] <= 0.01){
    presCluster1[i] = paste0(round(clus1Surp[i],4)," ***")
  } else if(clus1Surp[i] <= 0.05){
    presCluster1[i] = paste0(round(clus1Surp[i],4)," **")
  } else if(clus1Surp[i] <= 0.1){
    presCluster1[i] = paste0(round(clus1Surp[i],4)," *")
  } else {
    presCluster1[i] = as.character(round(clus1Surp[i],4))
  }
  if(clus2Surp[i] < 0.0001){
    presCluster2[i] = "< 0.0001 ***"
  } else if(clus2Surp[i] <= 0.01){
    presCluster2[i] = paste0(round(clus2Surp[i],4)," ***")
  } else if(clus2Surp[i] <= 0.05){
    presCluster2[i] = paste0(round(clus2Surp[i],4)," **")
  } else if(clus2Surp[i] <= 0.1){
    presCluster2[i] = paste0(round(clus2Surp[i],4)," *")
  } else {
    presCluster2[i] = as.character(round(clus2Surp[i],4))
  }
}
ABDF = data.frame(Sample = wineData$Sample[21:30], TheorisedCluster = c("Cluster 1",rep("Cluster 2",5),rep("Cluster 1",4)), Cluster1 = presCluster1, Cluster2 = presCluster2)
pander(ABDF)
```

```{r}
#summary table of PCs
PCs = prcomp(wineDataVariables, center = TRUE, scale = TRUE)
summ = summary(PCs)
SDs = summ$importance[1,]
cumVar = summ$importance[3,]
PCtable = data.frame(PC = colnames(summ$importance)[1:10], SD = SDs[1:10], Cvar = cumVar[1:10])
rownames(PCtable) = NULL
colnames(PCtable)[3] = "Cumulative Variance"
pander(PCtable)
```

```{r}
#table of surprising observations by model: full, cultivar split
sum(surprise<=.01) ; sum(surprise<=.05) ; sum(surprise<=.1)
surps = function(x){
  sSurp=sum((x<0.1 & x>=0.05))
  Surp = sum(x<0.05 & x>=.01)
  vSurp = sum(x<0.01)
  return(c(sSurp,Surp,vSurp))
}
surps(surprise) ; round(surprise,3)
# data.frame(Model = c("No Split","Cultivar Split"), SlSurp = c(surp(surprise)[1],surp(surpriseMencia[1:20])[1]+surp(surpriseAlicante[21:30])[1]),
#            Sur = c(surp(surprise)[2],
#                    surp(surpriseMencia[1:20])[2] + surp(surpriseAlicante[21:30])[2]),
#            vsur = c(surp(surprise)))

modMat = matrix(c(surps(surprise),surps(surpriseMencia[1:20])+surps(surpriseAlicante[21:30])),
                nrow=2,
                ncol=3,
                byrow=TRUE)

modMat
surObs = as.data.frame(modMat, row.names = c("Full Model","Cultivar Split"))
colnames(surObs) = c("Slightly Surprising", "Surprising", "Very Surprising")
surObs
```


Writeup

Materials

  Data Assembly

    Data were extracted from {revilla et al} using the docxtractor package {citation} in R {version} {citation}, following conversion of the available .pdf file to a .docx file using Microsoft Office 365 {citation} .

  Statistical Analyses
  
    All statistical analyses were carried out using the base or stats packages of R {version} {citation}.
    
  Plots and Tables
  
    Plots were created using the ggplot2 {+base +others} package {citation} in R {version} {citation} . Tables were created using the pander {/knitr} package {citation}.
    
Methods
  
  Data Assembly {already in report}
  
  Principal Component Analysis
  
    Principal Component Analysis (PCA) {citation?} was performed on the dataset extracted from {revilla et al} for the purpose of dimensionality reduction. Principal Components were derived from the full dataset, not from subgroupings of observations. Principal Components with retained variance greater than 1 were kept for use in analysis, and the rest discarded.
    
  Normality Assessment
    
    Normality of Principal Components, both in aggregate and within relevant subgroupings, was assessed using Shapiro-Wilk tests {citation?} .
    
  Graphical Assessment
  
    A scatterplot of Principal Component 1 (PC1) against Principal Component 2 (PC2) was plotted and assessed visually for separation between the Mencia and Alicante Bouschet cultivars, as well as any other apparent subgroupings.
    
  Surprising Observations Analysis
  
    Each observation's Mahalanobis distance {citation?} was calculated from the full dataset, from each cultivar, and from its own cultivar excluding itself. A p-value was calculated for how consistent each observation is with each of these distributions, using the property that the squared Mahalanobis distance of points in a shared multivariate normal distribution follows a {chi-squared M} distribution, where M is the number of variables being considered {citation}. Observations with a p-value less than .1 were labelled as "Slightly Surprising," those with p < .05 were labelled "Surprising," and those with p < .01 were labelled "Very Surprising."
  
  Subgrouping Analysis
  
    Following the identification of a potential subgrouping within the Alicante Bouschet cultivar, the Graphical Assessment and Surprising Observation Analysis methods were applied to the Alicante Bouschet samples, with samples T22 to T26 categorised as "Cluster 2" and the rest as "Cluster 1." PC4 and PC5 needed to be excluded from this analysis, as the Mahalanobis distance cannot be meaningfully calculated unless the number of datapoints used to calculate the covariance matrix exceeds the number of variables being considered by at least 2 {citation}. For each sample, the Mahalanobis distance was calculated from each theorised cluster, and the p-value obtained.
    
    
Results Writeup
  
  Principal Component Analysis
  
    A Scree plot of the Principal Components (PCs) showed possible elbow points at PC2 and PC4, while the first five PCs each retained variance greater than 1 {figure scree}. We retained the first five PCs, which collectively retained 90.9% of the sample variance {table PCs}.
    
  Normality Assessment
  
    Shapiro-Wilk tests on the derived PCs showed sufficient evidence to claim that PC1 (p=.0001) and PC5 (p=.01) were not normally distributed across the overall dataset. When assessed within each cultivar, both of these PCs appeared to follow a normal distribution, but PC2 did not appear to show a normal distribution within the Mencia (p=.0161) or Alicante Bouschet (p=.0095) cultivars {table Shapiro}.
    
  Graphical Assessment
  
    PC1 and PC2 for the whole dataset are plotted on {table fullscatter}. The separation by cultivar observed by {revilla et al} is clearly replicated here. Samples M16 and M17 are noted as probable outliers within the Mencia cultivar, but there does not appear to be any indication of subgroupings within the Mencia samples. The Alicante Bouschet Samples, however, show a clear separation, with four samples diverging from the rest along PC2. Sample T22 also appears to be an outlier relative to the comparatively very tight clustering of T21 and T27-T30.
    
  Surprising Observations Analysis
  
    Under the assumption that all samples share an underlying multivariate normal distribution, sample M8 is a significant outlier (p=.0096), while samples M17 (p=.0747) and T23 (p=.0987) are slightly but not severely deviant from the assumed distribution {table mahalfull}. When each cultivar is assumed to form a separate multivariate normal distribution, however, there are no strong outliers and three slightly surprising samples: M7 (p=.0732), M8  (p=.0583), and M17(p=.0579) {table mahalfull}. This lends moderate additional evidence to the claim that the variables measured by {revilla et al} can reliably separate wine samples of these two cultivars. More compellingly, we see that every sample is highly inconsistent with the distribution of the opposing cultivar (p=.0057 for sample T25, p<.0001 for all other samples) {table mahalfull}, hence it is very unlikely that samples of either of these cultivars could be mistakenly categorised as belonging to the other cultivar when these measurements are known.