---
title: "Group report"
format:
  typst:
    number-sections: true
    fig-format: png
editor: visual
bibliography: references.bib
execute:
  warning: false
  message: false
  echo: false
---

## Introduction {#sec-introduction}

The phenolic composition of red wine particularly its anthocyanin and tannin content governs colour, astringency, bitterness and ageing potential, and it is shaped jointly by grape cultivar, growing conditions and winemaking practice. Because anthocyanin composition is under strong genetic control, the relative proportions of individual anthocyanins (the "anthocyanin fingerprint") have been proposed as an analytical tool for verifying the grape cultivar used to produce a wine, complementing more general spectrophotometric measures of colour and phenolic content.

@revilla2016 studied this question for wines from the Appellation d'Origine Contrôlée (AOC) Valdeorras region (Galicia, NW Spain), comparing single-cultivar young red wines made from Mencía the principal red cultivar of the region and Alicante Bouschet, a teinturier cultivar with red-pigmented flesh that is often blended with Mencía to boost colour. Thirty wines from the 2014 vintage (20 Mencía, 10 Alicante Bouschet) were analysed after malolactic fermentation using two complementary techniques: UV-Vis spectrophotometry, to quantify total phenols, colour intensity, hue, chemical age and tannin- and anthocyanin-related indices; and HPLC-DAD, to determine the relative proportions of 14 individual anthocyanins (5 monoglucosides and 9 acylated derivatives). The authors used principal component analysis and linear discriminant analysis on each variable block separately and found that both effectively separated the two cultivars.

For this project we combine the quantitative data reported in Tables 1, 2, 5 and 6 of @revilla2016 into a single data frame of 30 wines and 23 numeric variables (5 spectrophotometric/colour variables, 4 anthocyanin/tannin concentration variables, and 14 relative-proportion anthocyanin variables), together with the grape cultivar (Mencía or Alicante Bouschet) as a grouping variable. Unlike the original paper, which analysed the spectrophotometric and HPLC variable blocks separately, we treat all 23 variables jointly, which lets us ask whether the wines separate as cleanly on the full phenolic profile as they did on either block alone.

Our aims are to describe this dataset quantitatively and graphically, assess whether a multivariate Gaussian model is a reasonable approximation for the joint distribution of these variables, identify any atypical wines, and determine whether unsupervised grouping of the samples recovers the Mencía/Alicante Bouschet distinction and, if not, to discuss why.

The remainder of the report is organised as follows. @sec-methodology describes how the data frame was assembled and the statistical methods used. @sec-results presents the descriptive, graphical and multivariate analyses. @sec-discussion interprets the findings in light of @revilla2016's own results and discusses their implications.

## Methodology {#sec-methodology}

### Materials

#### Data Assembly

Data were extracted from a .docx copy of @revilla2016 using the docxtractor package @docxtractr in R version 4.4.2 @Rlanguage.

#### Statistical Analyses

All statistical analyses were carried out using the base or stats packages of R version 4.4.2 @Rlanguage.

#### Plots and Tables \[unfinished\]

Plots were created using the ggplot2 {+base +others} package {citation} in R {version} {citation} . Tables were created using the pander {/knitr} package {citation}.

### Methods

### Data assembly

We combined the quantitative data reported in Tables 1, 2, 5 and 6 of @revilla2016 into a single data frame, using the sample labelling from the original tables (`M` = Mencía, `T` = Alicante Bouschet) to derive a `Cultivar` grouping variable.

```{r}
#| label: read-data
#| echo: true
library(tidyverse)

wine <- read_csv("extracted data.csv") |>
  select(-`...1`) |>
  mutate(
    Cultivar = factor(if_else(str_starts(Sample, "M"), "Mencia", "Alicante Bouschet")),
    .after = Sample
  )
```

The resulting data frame contains 30 wines (20 Mencía, 10 Alicante Bouschet) and 23 numeric phenolic variables: 5 spectrophotometric/colour variables (`Tot_Phen_Ind`, `Col_Int_420_520`, `Col_Int_420_520_620`, `Hue`, `Chem_Age_Ind`) and 4 anthocyanin/tannin concentration variables (`Tot_Anthocyans`, `Tot_Anthocyanins`, `Col_Anthocyanins`, `Tot_Tannins`) from Tables 1 and 2, and 14 relative-proportion anthocyanin variables — 5 monoglucosides and 9 acylated derivatives (`DpGl` through `MvGlCf`) — from Tables 5 and 6.

#### Principal Component Analysis \[unfinished\]

Principal Component Analysis (PCA) {citation?} was performed on the dataset extracted from @revilla2016 for the purpose of dimensionality reduction. Principal Components were derived from the full dataset, not from subgroupings of observations. Principal Components with retained variance greater than 1 were kept for use in analysis, and the rest discarded.

#### Normality Assessment \[unfinished\]

Normality of Principal Components, both in aggregate and within relevant subgroupings, was assessed using Shapiro-Wilk tests {citation?} .

#### Graphical Assessment

A scatterplot of Principal Component 1 (PC1) against Principal Component 2 (PC2) was plotted and assessed visually for separation between the Mencia and Alicante Bouschet cultivars, as well as any other apparent subgroupings.

#### Surprising Observations Analysis

Each observation's Mahalanobis distance {citation?} was calculated from the full dataset, from each cultivar, and from its own cultivar excluding itself. A p-value was calculated for how consistent each observation is with each of these distributions, using the property that the squared Mahalanobis distance of points in a shared multivariate normal distribution follows a {chi-squared M} distribution, where M is the number of variables being considered {citation}. Observations with a p-value less than .1 were labelled as "Slightly Surprising," those with p \< .05 were labelled "Surprising," and those with p \< .01 were labelled "Very Surprising."

#### Subgrouping Analysis \[unfinished\]

Following the identification of a potential subgrouping within the Alicante Bouschet cultivar, the Graphical Assessment and Surprising Observation Analysis methods were applied to the Alicante Bouschet samples, with samples T22 to T26 categorised as "Cluster 2" and the rest as "Cluster 1." PC4 and PC5 needed to be excluded from this analysis, as the Mahalanobis distance cannot be meaningfully calculated unless the number of datapoints used to calculate the covariance matrix exceeds the number of variables being considered by at least 2 {citation}. For each sample, the Mahalanobis distance was calculated from each theorised cluster, and the p-value obtained.

### Statistical methods

This Report makes use of Principle Component Analysis to .... We also make use of the Shapiro-Wilk test of Normality to judge whether are variables are normally distributed on a univariate level.

## Results {#sec-results}

### Descriptive analysis {#sec-descriptive}

```{r}
#| label: tbl-summary-colour
#| tbl-cap: "Mean, standard deviation, and range of the spectrophotometric and colour variables by grape cultivar."

wine |>
  select(Cultivar, Tot_Phen_Ind, Col_Int_420_520, Col_Int_420_520_620, Hue, Chem_Age_Ind) |>
  pivot_longer(-Cultivar, names_to = "Variable", values_to = "value") |>
  summarise(
    Mean = mean(value), SD = sd(value), Min = min(value), Max = max(value),
    .by = c(Cultivar, Variable)
  ) |>
  arrange(Variable, Cultivar) |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  knitr::kable()
```

The two cultivars differ sharply on pigment-related colour measures. Alicante Bouschet wines have roughly twice the colour intensity of Mencía wines on both scales (means of 19.07 versus 8.47, and 21.44 versus 10.60) and close to twice the total phenol index (65.55 versus 36.25), consistent with the teinturier character of the cultivar: because the berry flesh is pigmented as well as the skin, pressing yields an already-coloured must. The two maturation indices behave differently. Chemical age is statistically indistinguishable between the groups (0.10 versus 0.09) and hue is only slightly lower in Alicante Bouschet, with Mencía spanning a considerably wider range (0.43–0.69 against 0.49–0.60). The cultivar difference is therefore one of pigment quantity rather than of maturation state. It is worth noting that despite the large mean separations, none of these variables separates the cultivars completely: the most deeply coloured Mencía wines exceed the palest Alicante Bouschet wines on every measure.

```{r}
#| label: tbl-summary-concentration
#| tbl-cap: "Mean, standard deviation, and range of the anthocyanin and tannin concentration variables (mg/L) by grape cultivar."

wine |>
  select(Cultivar, Tot_Anthocyans, Tot_Anthocyanins, Col_Anthocyanins, Tot_Tannins) |>
  pivot_longer(-Cultivar, names_to = "Variable", values_to = "value") |>
  summarise(
    Mean = mean(value), SD = sd(value), Min = min(value), Max = max(value),
    .by = c(Cultivar, Variable)
  ) |>
  arrange(Variable, Cultivar) |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  knitr::kable()
```

The concentration measures reinforce this pattern, with Alicante Bouschet wines containing roughly two to three times the total anthocyans, total anthocyanins, coloured anthocyanins and total tannins of Mencía wines. Two features stand out. First, the Alicante Bouschet group is markedly more heterogeneous, with standard deviations two to three times those of the Mencía group on every variable; @revilla2016 attribute this to deliberate variation in maceration length, since these wines are commonly produced for blending and are managed to different phenolic targets. Second, the ranges overlap substantially in every case — most strikingly for total anthocyans, where the Mencía maximum of 892 mg/L exceeds the Alicante Bouschet minimum of 438 mg/L. This overlap is driven by two unusual Mencía wines, M16 and M17, which combine high anthocyanin content with the lowest tannin levels in the dataset; these observations recur in the analysis of atypical wines in @sec-outliers.

```{r}
#| label: tbl-summary-monoglucosides
#| tbl-cap: "Mean, standard deviation, and range of the relative content (%) of anthocyanidin monoglucosides by grape cultivar."

wine |>
  select(Cultivar, DpGl, CyGl, PtGl, PnGl, MvGl) |>
  pivot_longer(-Cultivar, names_to = "Variable", values_to = "value") |>
  summarise(
    Mean = mean(value), SD = sd(value), Min = min(value), Max = max(value),
    .by = c(Cultivar, Variable)
  ) |>
  arrange(Variable, Cultivar) |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  knitr::kable()
```

The relative proportions of the anthocyanidin monoglucosides provide the cleanest cultivar distinction in the dataset. Malvidin-3-O-glucoside dominates in both groups but accounts for a substantially higher share in Mencía wines (59.17% against 47.08%), and the difference is driven by peonidin-3-O-glucoside, present at more than three times the relative level in Alicante Bouschet (26.28% against 8.10%). Its biosynthetic precursor cyanidin-3-O-glucoside follows the same pattern at lower absolute levels. Unlike any of the variables considered so far, the ranges of these three compounds do not overlap between cultivars at all, making them the most discriminating single measurements available. Delphinidin-3-O-glucoside is the exception: its group means are essentially identical (4.44% and 4.50%), so it carries no cultivar information despite belonging to the same compound class.

```{r}
#| label: tbl-summary-acylated
#| tbl-cap: "Mean, standard deviation, and range of the relative content (%) of acylated anthocyanins by grape cultivar."

wine |>
  select(Cultivar, DpGlAc, PtGlAc, PnGlAc, MvGlAc, DpGlCm, PtGlCm, PnGlCm, MvGlCm, MvGlCf) |>
  pivot_longer(-Cultivar, names_to = "Variable", values_to = "value") |>
  summarise(
    Mean = mean(value), SD = sd(value), Min = min(value), Max = max(value),
    .by = c(Cultivar, Variable)
  ) |>
  arrange(Variable, Cultivar) |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  knitr::kable()
```

Acylation distinguishes the cultivars through its pattern rather than its extent. All four acetylated derivatives are more abundant in Mencía wines, with the acetyl derivative of malvidin-3-O-glucoside showing the largest gap (8.74% against 3.36%) and, together with the acetyl derivative of petunidin, forming a further pair of non-overlapping ranges. The coumaroylated derivatives trend in the opposite direction, though less consistently: the delphinidin and peonidin forms are clearly higher in Alicante Bouschet, while the malvidin and petunidin forms differ little between groups. The caffeoyl derivative of malvidin shows no cultivar difference. This reversal in acylation preference, rather than any difference in total acylation, is what carries the varietal signal.

### Graphical analysis {#sec-graphical}

```{r}
#| label: fig-correlation-heatmap
#| fig-cap: "Pairwise Pearson correlations among the 23 phenolic variables."
#| fig-width: 9
#| fig-height: 8

library(ggcorrplot)

wine |>
  select(-Sample, -Cultivar) |>
  cor() |>
  ggcorrplot(
    type = "lower",
    method = "circle",
    lab = TRUE,
    lab_size = 2,
    hc.order = TRUE,
    colors = c("blue", "white", "red"),
    legend.title = "Corr"
  )
```

COULD POTENTIALLY CUT THIS KIND OF REDUNDANT /

```{r Univariate Distributions}
#| label: fig-Univariate-distributions
#| fig-cap: "Univariate Distributions, by Cultivar"

wine |>
  pivot_longer(where(is.numeric), names_to = "Variable", values_to = "value") |>
  ggplot(aes(value, fill = Cultivar)) +
  geom_histogram(aes(y = after_stat(density)), alpha = 0.5, bins = 12, position = "identity") +
  facet_wrap(~Variable, scales = "free") +
  theme_minimal(base_size = 10)
```

histograms / density plots with normal overlay

correlation heat map --\> can help show why we need PCA.

### Assessing multivariate normality {#sec-normality}

PLAN: MVN means Univariate normality per variable (testing each variables normality) since we only have 30 observations (n \< 50) Ill use shapiro-wilk for normality

```{r}
#| label: tbl-shapiro-univariate
#| tbl-cap: "Shapiro-Wilk tests of univariate normality for each phenolic variable."

library(rstatix)

wine |>
  group_by(Cultivar) |>
  shapiro_test(
    Tot_Phen_Ind, Col_Int_420_520, Col_Int_420_520_620, Hue, Chem_Age_Ind,
    Tot_Anthocyans, Tot_Anthocyanins, Col_Anthocyanins, Tot_Tannins,
    DpGl, CyGl, PtGl, PnGl, MvGl,
    DpGlAc, PtGlAc, PnGlAc, MvGlAc,
    DpGlCm, PtGlCm, PnGlCm
  ) |>
  mutate(
    statistic = round(statistic, 3),
    p = round(p, 5)
  ) |>
  rename(Variable = variable, W = statistic, `p-value` = p) |>
  arrange(`p-value`) |>
  knitr::kable()
```

```{r}

```

Splitting by cultivar clears up some of pooled sample results for "Tot_Anthocyanins" and "Tot_Anthocyans", both strongly non-normal pooled with extremely small p-values, but are fine within the Alicante Bouschet Cultivar. This suggests a gap from @sec-descriptive creating an artificial mixture effect, not real non-normality

12 of 42 within-cultivar tests still reject at 5% significance, this is too many to be random as with 42 variables we would expect about 2 to reject by chance (42 \* 0.05 = 2.1). Mencia's total concentration variables are very poor "Tot_Anthocyanins", "Tot_Anthocyans" and coumaroylated derivatives are rejected for both cultivars "DpGICm" and "PnGICm" also in just mencia "PtGICm" which is already flagged before as the least consistent acylated compounds.

Since marginal normality is necessary for MVN, these rejection are enough to conclude neither cultivars phenolic profile is multivariate normal. However, we must consider a power caveat where the probability we reject $H_0$ given $H_1$ is true is dictated by the number of observations and alicante Bouschet has n=10 observations which half the size of mencia. So, when we find this cultivar has fewer rejections it may not just indicate its more normal but that its harder to detect non-normality.

```{r}
#| label: fig-flagged-distributions
#| fig-cap: "Distributions of variables rejecting univariate normality, by cultivar."
flagged <- c("Tot_Anthocyanins", "Tot_Anthocyans", "Tot_Tannins",
             "DpGlCm", "PnGlCm", "PtGlCm")

wine |>
  select(Cultivar, all_of(flagged)) |>
  pivot_longer(-Cultivar, names_to = "Variable", values_to = "value") |>
  ggplot(aes(value, fill = Cultivar)) +
  geom_histogram(aes(y = after_stat(density)), alpha = 0.5, bins = 12, position = "identity") +
  facet_wrap(~Variable, scales = "free") +
  theme_minimal(base_size = 11) +
  labs(title = "Distributions of Variables Rejecting Univariate Normality")
```

### Atypical observations {#sec-outliers}

The atypicality of a datapoint relative to a group distribution is measured by the Mahalanobis distance. For this dataset, atypicality will be assessed based on distance from the distribution of the cultivar samples, not from the overall dataset. Unfortunately, the number of variables measured, 23, exceeds the sample size of either cultivar, resulting in a singular, non-invertible covariance matrix, making the calculation of the Mahalanobis distance impossible. As such, we will use Principal Component Analysis and remove the non-

```{r}
#| label: tbl-mahalanobis-pooled
#| tbl-cap: "Mahalanobis distances and associated tail probabilities for all 30 wines, computed on the full set of 23 phenolic variables."

wineData <- as.data.frame(wine)
wineDataVariables <- wineData[, 3:25]

wineData$mahalanobis <- mahalanobis(
  wineDataVariables,
  center = colMeans(wineDataVariables),
  cov = cov(wineDataVariables)
)
wineData$surprise <- pchisq(wineData$mahalanobis, df = 23, lower.tail = FALSE)

wineData |>
  select(Sample, Cultivar, mahalanobis, surprise) |>
  mutate(
    mahalanobis = round(mahalanobis, 2),
    surprise = round(surprise, 4)
  ) |>
  rename(Wine = Sample, `Mahalanobis distance` = mahalanobis, `Tail probability` = surprise) |>
  arrange(desc(`Mahalanobis distance`)) |>
  knitr::kable()
```

```{r}
#| label: tbl-shapiro-pca
#| tbl-cap: "Shapiro-Wilk tests of normality applied to each principal component."

winePCA <- prcomp(wineDataVariables, center = TRUE, scale = TRUE)
wineDataPCADF <- as.data.frame(winePCA$x)

tibble(
  Component = colnames(wineDataPCADF),
  W = sapply(wineDataPCADF, \(x) shapiro.test(x)$statistic),
  `p-value` = sapply(wineDataPCADF, \(x) shapiro.test(x)$p.value)
) |>
  mutate(W = round(W, 3), `p-value` = round(`p-value`, 5)) |>
  knitr::kable()
```

```{r}
#| label: tbl-group-rank
#| tbl-cap: "Sample size, number of variables, and covariance matrix rank for each cultivar."

menciaData   <- wineData[wineData$Cultivar == "Mencia", 3:25]
alicanteData <- wineData[wineData$Cultivar == "Alicante Bouschet", 3:25]

tibble(
  Cultivar = c("Mencia", "Alicante Bouschet"),
  n = c(nrow(menciaData), nrow(alicanteData)),
  p = c(ncol(menciaData), ncol(alicanteData)),
  `Max rank (n-1)` = n - 1,
  `Covariance rank` = c(qr(cov(menciaData))$rank, qr(cov(alicanteData))$rank)
) |>
  knitr::kable()
```

Within-cultivar Mahalanobis distances cannot be computed on the full variable set. The covariance matrix of a group of $n$ observations has rank at most $n-1$, so with 20 Mencía and 10 Alicante Bouschet wines against 23 variables, both within-group covariance matrices are singular and cannot be inverted (@tbl-group-rank). We therefore restrict the within-group analysis to the leading principal components of each cultivar, retaining enough components that the covariance matrix is of full rank.

```{r}
#| label: fig-scree-groups
#| fig-cap: "Proportion of variance explained by each principal component, computed separately within each cultivar."
#| fig-width: 7
#| fig-height: 4

scree_data <- function(dat, cultivar) {
  p <- prcomp(dat, center = TRUE, scale. = TRUE)
  tibble(
    Cultivar = cultivar,
    PC = seq_along(p$sdev),
    Proportion = p$sdev^2 / sum(p$sdev^2)
  )
}

bind_rows(
  scree_data(menciaData, "Mencia"),
  scree_data(alicanteData, "Alicante Bouschet")
) |>
  filter(PC <= 10) |>
  ggplot(aes(PC, Proportion)) +
  geom_col(fill = "grey40") +
  facet_wrap(~Cultivar) +
  scale_x_continuous(breaks = 1:10) +
  labs(x = "Principal component", y = "Proportion of variance explained") +
  theme_minimal()
```

```{r}
#| label: tbl-mahalanobis-groupwise
#| tbl-cap: "Within-cultivar Mahalanobis distances computed on the leading principal components of each group."

pca_mahal <- function(dat, k) {
  scores <- prcomp(dat, center = TRUE, scale. = TRUE)$x[, 1:k, drop = FALSE]
  mahalanobis(scores, center = colMeans(scores), cov = cov(scores))
}

bind_rows(
  tibble(
    Wine = wineData$Sample[wineData$Cultivar == "Mencia"],
    Cultivar = "Mencia",
    Distance = pca_mahal(menciaData, 2)
  ),
  tibble(
    Wine = wineData$Sample[wineData$Cultivar == "Alicante Bouschet"],
    Cultivar = "Alicante Bouschet",
    Distance = pca_mahal(alicanteData, 3)
  )
) |>
  mutate(Distance = round(Distance, 2)) |>
  arrange(Cultivar, desc(Distance)) |>
  knitr::kable()
```

```{r}
# PCA
PCs = prcomp(wineDataVariables, center = TRUE, scale = TRUE)
screeplot(PCs,type = "lines") +
  abline(h=1, col="red")
```

```{r}
# Shapiro-Wilk test of PC1-5 aggregate and by cultivar
library(pander)
aggShap = rep(NA,5)
menShap = rep(NA,5)
abShap = rep(NA,5)
for(i in 1:5){
  aggShap[i] = shapiro.test(wineDataPCADF[,i])$p.value
  menShap[i] = shapiro.test(wineDataPCADF[1:20,i])$p.value
  abShap[i] = shapiro.test(wineDataPCADF[21:30,i])$p.value
}
shapDF = data.frame(Component = c("PC1","PC2","PC3","PC4","PC5"), Aggregate = aggShap, Mencia = menShap, AlicanteBouschet = abShap)
colnames(shapDF)[4] = "Alicante Bouschet"

for(i in 1:5){
  for(j in 2:4){
    charTemp = as.character(round(as.numeric(shapDF[i,j]),4))
    shapDF[i,j] = ifelse(shapDF[i,j] <= 0.01, paste0(charTemp," ***"),
                        ifelse(shapDF[i,j] <= 0.05, paste0(charTemp, " **"),
                               ifelse(shapDF[i,j] <=0.1, paste0(charTemp, " *"),
                                      charTemp)))
  }
}
pander(shapDF)
```

```{r}
# graphical of PCs, full dataset
plot(wineDataPCADF$PC1,
     wineDataPCADF$PC2,
     col = c(rep("red",20),rep("blue",10)),
     lwd = 3) + text(wineDataPCADF$PC1+0.1, wineDataPCADF$PC2-0.2, wineData$Sample, cex = 0.5)
```

```{r}
# surprising results for mahalanobis: full model, cultivar split, cultivar split with exclusion
presFullModel = rep(NA,30)
presMenInc = rep(NA,30)
presABInc = rep(NA,30)
presSplitExc = rep(NA,30)
wineDataPCADFR = wineDataPCADF[,1:5]
PCAMahalanobis = mahalanobis(wineDataPCADFR,center = colMeans(wineDataPCADFR),cov = cov(wineDataPCADFR))
surprise = pchisq(PCAMahalanobis,df=5,lower.tail = FALSE)
PCAFullMenciaMahal = mahalanobis(wineDataPCADFR, center=colMeans(wineDataPCADFR[1:20,]), cov=cov(wineDataPCADFR[1:20,]))
surpriseMencia = pchisq(PCAFullMenciaMahal, df=5, lower.tail=FALSE)
PCAFullAlicanteMahal = mahalanobis(wineDataPCADFR, center=colMeans(wineDataPCADFR[21:30,]), cov=cov(wineDataPCADFR[21:30,]))
surpriseAlicante = pchisq(PCAFullAlicanteMahal, df=5, lower.tail=FALSE)
surpriseCulExc = rep(NA,30)
mahalCulExc = rep(NA,30)
for(i in 1:20){
  wineTemp = wineDataPCADFR[1:20,][-i,]
  muTemp = colMeans(wineTemp)
  covTemp = cov(wineTemp)
  mahalCulExc[i] = mahalanobis(wineDataPCADFR[i,], center = muTemp, cov=covTemp)
  surpriseCulExc[i] = pchisq(mahalCulExc[i], df=5, lower.tail=FALSE)
}
for(i in 21:30){
  iAdj = i - 20
  wineTemp = wineDataPCADFR[21:30,][-iAdj,]
  muTemp = colMeans(wineTemp)
  covTemp = cov(wineTemp)
  mahalCulExc[i] = mahalanobis(wineDataPCADFR[i,], center = muTemp, cov=covTemp)
  surpriseCulExc[i] = pchisq(mahalCulExc[i], df=5, lower.tail=FALSE)
}

for(i in 1:30){
  if(surprise[i] < 0.0001){
    presFullModel[i] = "< 0.0001 ***"
  } else if(surprise[i] <= 0.01){
    presFullModel[i] = paste0(round(surprise[i],4)," ***")
  } else if(surprise[i] <= 0.05){
    presFullModel[i] = paste0(round(surprise[i],4)," **")
  } else if(surprise[i] <= 0.1){
    presFullModel[i] = paste0(round(surprise[i],4)," *")
  } else {
    presFullModel[i] = as.character(round(surprise[i],4))
  }
  if(surpriseMencia[i] < 0.0001){
    presMenInc[i] = "< 0.0001 ***"
  } else if(surpriseMencia[i] <= 0.01){
    presMenInc[i] = paste0(round(surpriseMencia[i],4)," ***")
  } else if(surpriseMencia[i] <= 0.05){
    presMenInc[i] = paste0(round(surpriseMencia[i],4)," **")
  } else if(surpriseMencia[i] <= 0.1){
    presMenInc[i] = paste0(round(surpriseMencia[i],4)," *")
  } else {
    presMenInc[i] = as.character(round(surpriseMencia[i],4))
  }
  if(surpriseAlicante[i] < 0.0001){
    presABInc[i] = "< 0.0001 ***"
  } else if(surpriseAlicante[i] <= 0.01){
    presABInc[i] = paste0(round(surpriseAlicante[i],4)," ***")
  } else if(surpriseAlicante[i] <= 0.05){
    presABInc[i] = paste0(round(surpriseAlicante[i],4)," **")
  } else if(surpriseAlicante[i] <= 0.1){
    presABInc[i] = paste0(round(surpriseAlicante[i],4)," *")
  } else {
    presABInc[i] = as.character(round(surpriseAlicante[i],4))
  }
  if(surpriseCulExc[i] < 0.0001){
    presSplitExc[i] = "< 0.0001 ***"
  } else if(surpriseCulExc[i] <= 0.01){
    presSplitExc[i] = paste0(round(surpriseCulExc[i],4)," ***")
  } else if(surpriseCulExc[i] <= 0.05){
    presSplitExc[i] = paste0(round(surpriseCulExc[i],4)," **")
  } else if(surpriseCulExc[i] <= 0.1){
    presSplitExc[i] = paste0(round(surpriseCulExc[i],4)," *")
  } else {
    presSplitExc[i] = as.character(round(surpriseCulExc[i],4))
  }
}
mahalTable = data.frame(sample = wineData$Sample, fullModel = presFullModel, menciaInc = presMenInc, abInc = presABInc, splitExc = presSplitExc)
mahalTable$splitExc = gsub("2e-04","0.0002",mahalTable$splitExc)
colnames(mahalTable) = c("Sample","Full Dataset","Mencia","Alicante Bouschet","Cultivar Excluding Sample")
pander(mahalTable)
```

```{r}
# graph PC1-2 AB split
plot(wineDataPCADFR[21:30,]$PC1,
     wineDataPCADFR[21:30,]$PC2,
     col = c("green",rep("black",5),rep("green",4)),
     pch = as.character(c(1:9,0)))
```

```{r}
# mahalanobis AB PC1-3 without exclusion
presCluster1 = rep(NA,10)
presCluster2 = rep(NA,10)

wineAB1 = wineDataPCADFR[c(21,27:30),1:3]
wineAB2 = wineDataPCADFR[22:26,1:3]
muAB1 = colMeans(wineAB1)
muAB2 = colMeans(wineAB2)
sigAB1 = cov(wineAB1)
sigAB2 = cov(wineAB2)

clus1from1surp = pchisq(mahalanobis(wineAB1,center = muAB1, cov = sigAB1), df=2, lower.tail = FALSE)
clus2from1surp = pchisq(mahalanobis(wineAB2,center = muAB1, cov = sigAB1), df=2, lower.tail = FALSE)
clus2from2surp = pchisq(mahalanobis(wineAB2,center = muAB2, cov = sigAB2), df=2, lower.tail = FALSE)
clus1from2surp = pchisq(mahalanobis(wineAB1,center = muAB2, cov = sigAB2), df=2, lower.tail = FALSE)

clus1Surp = c(clus1from1surp[1],clus2from1surp,clus1from1surp[2:5])
clus2Surp = c(clus1from2surp[1],clus2from2surp,clus1from2surp[2:5])



for(i in 1:10){
  if(clus1Surp[i] < 0.0001){
    presCluster1[i] = "< 0.0001 ***"
  } else if(clus1Surp[i] <= 0.01){
    presCluster1[i] = paste0(round(clus1Surp[i],4)," ***")
  } else if(clus1Surp[i] <= 0.05){
    presCluster1[i] = paste0(round(clus1Surp[i],4)," **")
  } else if(clus1Surp[i] <= 0.1){
    presCluster1[i] = paste0(round(clus1Surp[i],4)," *")
  } else {
    presCluster1[i] = as.character(round(clus1Surp[i],4))
  }
  if(clus2Surp[i] < 0.0001){
    presCluster2[i] = "< 0.0001 ***"
  } else if(clus2Surp[i] <= 0.01){
    presCluster2[i] = paste0(round(clus2Surp[i],4)," ***")
  } else if(clus2Surp[i] <= 0.05){
    presCluster2[i] = paste0(round(clus2Surp[i],4)," **")
  } else if(clus2Surp[i] <= 0.1){
    presCluster2[i] = paste0(round(clus2Surp[i],4)," *")
  } else {
    presCluster2[i] = as.character(round(clus2Surp[i],4))
  }
}
ABDF = data.frame(Sample = wineData$Sample[21:30], Cluster1 = presCluster1, Cluster2 = presCluster2)
pander(ABDF)
```

### Grouping structure {#sec-clustering}

## Discussion {#sec-discussion}
