---
title: "Group report"
author:
  - name: "Riley Phillips"
  - name: "Wian de Vos"
  - name: "Oliver Temel"
  - name: "Simon Nelson"
format:
  typst:
    number-sections: true
    fig-format: png
editor: visual
bibliography: references.bib
execute:
  warning: false
  message: false
  echo: false
---

## Introduction {#sec-introduction}

The phenolic composition of red wine — particularly its anthocyanin and tannin content — governs colour, astringency, bitterness and ageing potential, and it is shaped jointly by grape cultivar, growing conditions and winemaking practice. Because anthocyanin composition is under strong genetic control, the relative proportions of individual anthocyanins (the "anthocyanin fingerprint") have been proposed as an analytical tool for verifying the grape cultivar used to produce a wine, complementing more general spectrophotometric measures of colour and phenolic content.

Revilla *et al.* studied this question for wines from the Appellation d'Origine Contrôlée (AOC) Valdeorras region (Galicia, NW Spain), comparing single-cultivar young red wines made from Mencía — the principal red cultivar of the region — and Alicante Bouschet, a teinturier cultivar with red-pigmented flesh that is often blended with Mencía to boost colour [@revilla2016]. Thirty wines from the 2014 vintage (20 Mencía, 10 Alicante Bouschet) were analysed after malolactic fermentation using two complementary techniques: UV-Vis spectrophotometry, to quantify total phenols, colour intensity, hue, chemical age and tannin- and anthocyanin-related indices; and HPLC-DAD, to determine the relative proportions of 14 individual anthocyanins (5 monoglucosides and 9 acylated derivatives). The authors used principal component analysis and linear discriminant analysis on each variable block separately and found that both effectively separated the two cultivars.

For this project we combine the quantitative data reported in Tables 1, 2, 5 and 6 of [@revilla2016] into a single data frame of 30 wines and 23 numeric variables (5 spectrophotometric/colour variables, 4 anthocyanin/tannin concentration variables, and 14 relative-proportion anthocyanin variables), together with the grape cultivar (Mencía or Alicante Bouschet) as a grouping variable. Unlike the original paper, which analysed the spectrophotometric and HPLC variable blocks separately, we treat all 23 variables jointly, which lets us ask whether the wines separate as cleanly on the full phenolic profile as they did on either block alone.

Our aims are to describe this dataset quantitatively and graphically, assess whether a multivariate Gaussian model is a reasonable approximation for the joint distribution of these variables, identify any atypical wines, and determine whether unsupervised grouping of the samples recovers the Mencía/Alicante Bouschet distinction — and, if not, to discuss why.

The remainder of the report is organised as follows. @sec-methodology describes how the data frame was assembled and the statistical methods used. @sec-results presents the descriptive, graphical and multivariate analyses. @sec-discussion interprets the findings in light of @revilla2016's own results and discusses their implications.

## Methodology {#sec-methodology}

### Materials

#### Data Assembly

We combined the quantitative data reported in Tables 1, 2, 5 and 6 of [@revilla2016] into a single data frame. Sample labels from the original tables were used to create a Cultivar grouping variable, with samples beginning with M classified as Mencía and those beginning with T classified as Alicante Bouschet.

The final dataset contained 30 wine samples, with 20 Mencía and 10 Alicante Bouschet wines, and 23 numerical variables. These included spectrophotometric and colour measurements, anthocyanin and tannin concentrations, and the relative proportions of individual anthocyanins.

Data were extracted from a .docx copy of [@revilla2016] using the docxtractor package [@docxtractr] in R version 4.4.2 [@Rlanguage].

#### Statistical Analyses

All statistical analyses were carried out in R version 4.4.2 [@Rlanguage]. Functions from the base and stats packages were used for correlation analysis, principal component analysis and Mahalanobis distance calculations. The rstatix package was used for the Shapiro-Wilk normality tests.

All statistical analyses were carried out using the base or stats packages of R version 4.4.2 [@Rlanguage].

#### Plots and Tables

Plots were created primarily using the ggplot2 package [@ggplot2] in R version 4.4.2 [@Rlanguage], with ggcorrplot [@ggcorrplot] used to produce the correlation heatmap. Tables were created using the \`kable()\` function from the knitr package [@knitr]. Data preparation and reshaping were carried out using functions from the tidyverse [@tidyverse].

### Methods

### Data assembly

We combined the quantitative data reported in Tables 1, 2, 5 and 6 of [@revilla2016] into a single data frame, using the sample labelling from the original tables (`M` = Mencía, `T` = Alicante Bouschet) to derive a `Cultivar` grouping variable.

```{r}
#| label: read-data
#| echo: true
library(tidyverse)

wine <- read_csv("extracted data.csv") |>
  select(-`...1`) |>
  mutate(
    Cultivar = factor(if_else(str_starts(Sample, "M"), "Mencia", "Alicante Bouschet")),
    .after = Sample
  )
```

The resulting data frame contains 30 wines (20 Mencía, 10 Alicante Bouschet) and 23 numeric phenolic variables: 5 spectrophotometric/colour variables (`Tot_Phen_Ind`, `Col_Int_420_520`, `Col_Int_420_520_620`, `Hue`, `Chem_Age_Ind`) and 4 anthocyanin/tannin concentration variables (`Tot_Anthocyans`, `Tot_Anthocyanins`, `Col_Anthocyanins`, `Tot_Tannins`) from Tables 1 and 2, and 14 relative-proportion anthocyanin variables — 5 monoglucosides and 9 acylated derivatives (`DpGl` through `MvGlCf`) — from Tables 5 and 6.

#### Principal Component Analysis

Principal Component Analysis (PCA) [@jolliffe2016] was performed on the dataset extracted from [@revilla2016] for the purpose of dimensionality reduction. Principal Components were derived from the full dataset, not from subgroupings of observations. Principal Components with retained variance greater than 1 were kept for use in analysis, and the rest discarded.

The proportion of variation explained by the principal components was examined using scree plots. Scores from the leading components were then used to look for patterns among the wine samples and determine whether Mencía and Alicante Bouschet naturally formed separate groups. Cultivar was not used to calculate the principal components, so any separation between the groups was based only on the measured phenolic characteristics.

#### Normality Assessment

The Shapiro-Wilk test was used to assess univariate normality for each numerical variable [@shapiro1965]. Tests were carried out separately for Mencía and Alicante Bouschet, as combining two groups with different distributions could make the pooled data appear non-normal simply because of the difference between cultivars.

A significance level of 0.05 was used, with a p-value below 0.05 treated as evidence against normality. Although these tests assess each variable separately rather than testing multivariate normality directly, univariate normality is a necessary condition of a multivariate normal distribution. Therefore, clear departures from normality across several variables were taken as evidence that multivariate normality may not be a suitable assumption. The smaller Alicante Bouschet sample size was also considered, as departures from normality are harder to detect with fewer observations.

#### Graphical Assessment

A scatterplot of Principal Component 1 (PC1) against Principal Component 2 (PC2) was plotted and assessed visually for separation between the Mencia and Alicante Bouschet cultivars, as well as any other apparent subgroupings.

#### Surprising Observations Analysis

Each observation's Mahalanobis distance was calculated from the full dataset, from each cultivar, and from its own cultivar excluding itself. A p-value was calculated for how consistent each observation is with each of these distributions, using the property that the squared Mahalanobis distance of points in a shared multivariate normal distribution follows a $\chi^2_m$ distribution, where $m$ is the number of variables being considered. Observations with a p-value less than .1 were labelled as "Slightly Surprising," those with p \< .05 were labelled "Surprising," and those with p \< .01 were labelled "Very Surprising."

#### Subgrouping Analysis

Following the identification of a potential subgrouping within the Alicante Bouschet cultivar, the Graphical Assessment and Surprising Observation Analysis methods were applied to the Alicante Bouschet samples, with samples T22 to T26 categorised as "Cluster 2" and the rest as "Cluster 1." PC4 and PC5 needed to be excluded from this analysis, as the Mahalanobis distance cannot be meaningfully calculated unless the number of datapoints used to calculate the covariance matrix exceeds the number of variables being considered by at least 2 [@pires2024] . For each sample, the Mahalanobis distance was calculated from each theorised cluster, and the p-value obtained.

## Results {#sec-results}

### Descriptive analysis {#sec-descriptive}

```{r}
#| label: tbl-summary-colour
#| tbl-cap: "Mean, standard deviation, and range of the spectrophotometric and colour variables by grape cultivar."

wine |>
  select(Cultivar, Tot_Phen_Ind, Col_Int_420_520, Col_Int_420_520_620, Hue, Chem_Age_Ind) |>
  pivot_longer(-Cultivar, names_to = "Variable", values_to = "value") |>
  summarise(
    Mean = mean(value), SD = sd(value), Min = min(value), Max = max(value),
    .by = c(Cultivar, Variable)
  ) |>
  arrange(Variable, Cultivar) |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  knitr::kable()
```

The two cultivars differ sharply on pigment-related colour measures. Alicante Bouschet wines have roughly twice the colour intensity of Mencía wines on both scales (means of 19.07 versus 8.47, and 21.44 versus 10.60) and close to twice the total phenol index (65.55 versus 36.25), consistent with the teinturier character of the cultivar: because the berry flesh is pigmented as well as the skin, pressing yields an already-coloured must. The two maturation indices behave differently. Chemical age is statistically indistinguishable between the groups (0.10 versus 0.09) and hue is only slightly lower in Alicante Bouschet, with Mencía spanning a considerably wider range (0.43–0.69 against 0.49–0.60). The cultivar difference is therefore one of pigment quantity rather than of maturation state. It is worth noting that despite the large mean separations, none of these variables separates the cultivars completely: the most deeply coloured Mencía wines exceed the palest Alicante Bouschet wines on every measure.

```{r}
#| label: tbl-summary-concentration
#| tbl-cap: "Mean, standard deviation, and range of the anthocyanin and tannin concentration variables (mg/L) by grape cultivar."

wine |>
  select(Cultivar, Tot_Anthocyans, Tot_Anthocyanins, Col_Anthocyanins, Tot_Tannins) |>
  pivot_longer(-Cultivar, names_to = "Variable", values_to = "value") |>
  summarise(
    Mean = mean(value), SD = sd(value), Min = min(value), Max = max(value),
    .by = c(Cultivar, Variable)
  ) |>
  arrange(Variable, Cultivar) |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  knitr::kable()
```

The concentration measures reinforce this pattern, with Alicante Bouschet wines containing roughly two to three times the total anthocyans, total anthocyanins, coloured anthocyanins and total tannins of Mencía wines. Two features stand out. First, the Alicante Bouschet group is markedly more heterogeneous, with standard deviations two to three times those of the Mencía group on every variable; Revilla *et al.* attribute this to deliberate variation in maceration length, since these wines are commonly produced for blending and are managed to different phenolic targets [@revilla2016]. Second, the ranges overlap substantially in every case — most strikingly for total anthocyans, where the Mencía maximum of 892 mg/L exceeds the Alicante Bouschet minimum of 438 mg/L. This overlap is driven by two unusual Mencía wines, M16 and M17, which combine high anthocyanin content with the lowest tannin levels in the dataset; these observations recur in the analysis of atypical wines in @sec-outliers.

```{r}
#| label: tbl-summary-monoglucosides
#| tbl-cap: "Mean, standard deviation, and range of the relative content (%) of anthocyanidin monoglucosides by grape cultivar."

wine |>
  select(Cultivar, DpGl, CyGl, PtGl, PnGl, MvGl) |>
  pivot_longer(-Cultivar, names_to = "Variable", values_to = "value") |>
  summarise(
    Mean = mean(value), SD = sd(value), Min = min(value), Max = max(value),
    .by = c(Cultivar, Variable)
  ) |>
  arrange(Variable, Cultivar) |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  knitr::kable()
```

The relative proportions of the anthocyanidin monoglucosides provide the cleanest cultivar distinction in the dataset. Malvidin-3-O-glucoside dominates in both groups but accounts for a substantially higher share in Mencía wines (59.17% against 47.08%), and the difference is driven by peonidin-3-O-glucoside, present at more than three times the relative level in Alicante Bouschet (26.28% against 8.10%). Its biosynthetic precursor cyanidin-3-O-glucoside follows the same pattern at lower absolute levels. Unlike any of the variables considered so far, the ranges of these three compounds do not overlap between cultivars at all, making them the most discriminating single measurements available. Delphinidin-3-O-glucoside is the exception: its group means are essentially identical (4.44% and 4.50%), so it carries no cultivar information despite belonging to the same compound class.

```{r}
#| label: tbl-summary-acylated
#| tbl-cap: "Mean, standard deviation, and range of the relative content (%) of acylated anthocyanins by grape cultivar."

wine |>
  select(Cultivar, DpGlAc, PtGlAc, PnGlAc, MvGlAc, DpGlCm, PtGlCm, PnGlCm, MvGlCm, MvGlCf) |>
  pivot_longer(-Cultivar, names_to = "Variable", values_to = "value") |>
  summarise(
    Mean = mean(value), SD = sd(value), Min = min(value), Max = max(value),
    .by = c(Cultivar, Variable)
  ) |>
  arrange(Variable, Cultivar) |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  knitr::kable()
```

Acylation distinguishes the cultivars through its pattern rather than its extent. All four acetylated derivatives are more abundant in Mencía wines, with the acetyl derivative of malvidin-3-O-glucoside showing the largest gap (8.74% against 3.36%) and, together with the acetyl derivative of petunidin, forming a further pair of non-overlapping ranges. The coumaroylated derivatives trend in the opposite direction, though less consistently: the delphinidin and peonidin forms are clearly higher in Alicante Bouschet, while the malvidin and petunidin forms differ little between groups. The caffeoyl derivative of malvidin shows no cultivar difference. This reversal in acylation preference, rather than any difference in total acylation, is what carries the varietal signal.

### Graphical analysis {#sec-graphical}

Having examined each variable as a uni-variate case in @sec-descriptive, the next logical step is considering how these 23 variables relate with each other. This matters decriptively as it answers questions like, do variables measuring related wine chemistry move together as we would expect? This has important methodologically consequences, since high correlation between variables limits the effective dimensional of the dataset and motivates the use of PCA or other dimensionality reduction.

```{r}
#| label: fig-correlation-heatmap
#| fig-cap: "Pairwise Pearson correlations among the 23 phenolic variables."
#| fig-width: 9
#| fig-height: 8

library(ggcorrplot)

wine |>
  select(-Sample, -Cultivar) |>
  cor() |>
  ggcorrplot(
    type = "lower",
    method = "circle",
    lab = TRUE,
    lab_size = 2,
    hc.order = TRUE,
    colors = c("blue", "white", "red"),
    legend.title = "Corr"
  )

# n_total <- nrow(cor_pairs)
# n_strong <- cor_pairs |> filter(abs(r) > 0.8) |> nrow()
# 
# n_strong
# n_strong / n_total
```

The correlation heatmap (@fig-correlation-heatmap) shows several near identical measurements. 'Tot_Anthocyanins' and 'Tot_Anthocyans' correlate at r = 0.998 and the colour/phenolic index variables ('Col_Int_420_520', 'Col_Int_420_520_620', 'Tot_Phen_Ind') all correlate at above 0.94 with one another. This is consistent with being derived from closely related measurements. The monoglucoside proportions also correlate with one another with an absolute r correlation coefficient of 0.95 for several pairs like 'MvGl' and 'PnGl' this is expected for pairs of compositional variables whose relative shares are constrained to sum to 100% of monoglucoside. There were 51 pairs out of all possible 2 variable combinations with an absolute value of r \> 0.8, which as a percentage is 20.16% (2dp) of all combinations. This is a relatively high percentage of total pairs and means there is substantial redundancy amongst the 23 variables.

### Assessing multivariate normality {#sec-normality}

```{r}
#| label: tbl-shapiro-univariate
#| tbl-cap: "Shapiro-Wilk tests of univariate normality for each phenolic variable."

library(rstatix)

wine |>
  group_by(Cultivar) |>
  shapiro_test(
    Tot_Phen_Ind, Col_Int_420_520, Col_Int_420_520_620, Hue, Chem_Age_Ind,
    Tot_Anthocyans, Tot_Anthocyanins, Col_Anthocyanins, Tot_Tannins,
    DpGl, CyGl, PtGl, PnGl, MvGl,
    DpGlAc, PtGlAc, PnGlAc, MvGlAc,
    DpGlCm, PtGlCm, PnGlCm
  ) |>
  mutate(
    statistic = round(statistic, 3),
    p = round(p, 5)
  ) |>
  rename(Variable = variable, W = statistic, `p-value` = p) |>
  arrange(`p-value`) |>
  knitr::kable()
```

Splitting by cultivar clears up some of pooled sample results for "Tot_Anthocyanins" and "Tot_Anthocyans", both strongly non-normal pooled with extremely small p-values, but are fine within the Alicante Bouschet Cultivar. This suggests a gap from @sec-descriptive creating an artificial mixture effect, not real non-normality

12 of 42 within-cultivar tests still reject at 5% significance, this is too many to be random as with 42 variables we would expect about 2 to reject by chance (42 \* 0.05 = 2.1). Mencia's total concentration variables are very poor "Tot_Anthocyanins", "Tot_Anthocyans" and coumaroylated derivatives are rejected for both cultivars "DpGICm" and "PnGICm" also in just mencia "PtGICm" which is already flagged before as the least consistent acylated compounds.

Since marginal normality is necessary for MVN, these rejection are enough to conclude neither cultivars phenolic profile is multivariate normal. However, we must consider a power caveat where the probability we reject $H_0$ given $H_1$ is true is dictated by the number of observations and alicante Bouschet has n=10 observations which half the size of mencia. So, when we find this cultivar has fewer rejections it may not just indicate its more normal but that its harder to detect non-normality.

```{r}
#| label: fig-flagged-distributions
#| fig-cap: "Distributions of variables rejecting univariate normality, by cultivar."
flagged <- c("Tot_Anthocyanins", "Tot_Anthocyans", "Tot_Tannins",
             "DpGlCm", "PnGlCm", "PtGlCm")

wine |>
  select(Cultivar, all_of(flagged)) |>
  pivot_longer(-Cultivar, names_to = "Variable", values_to = "value") |>
  ggplot(aes(value, fill = Cultivar)) +
  geom_histogram(aes(y = after_stat(density)), alpha = 0.5, bins = 12, position = "identity") +
  facet_wrap(~Variable, scales = "free") +
  theme_minimal(base_size = 11) +
  labs(title = "Distributions of Variables Rejecting Univariate Normality")
```

### Atypical observations {#sec-outliers}

The atypicality of a datapoint relative to a group distribution is measured by the Mahalanobis distance. For this dataset, atypicality will be assessed based on distance from the distribution of the cultivar samples, not from the overall dataset. Unfortunately, the number of variables measured, 23, exceeds the sample size of either cultivar, resulting in a singular, non-invertible covariance matrix, making the calculation of the Mahalanobis distance impossible. As such, we will use Principal Component Analysis and remove the non-

```{r}
#| label: tbl-mahalanobis-pooled
#| tbl-cap: "Mahalanobis distances and associated tail probabilities for all 30 wines, computed on the full set of 23 phenolic variables."

wineData <- as.data.frame(wine)
wineDataVariables <- wineData[, 3:25]

wineData$mahalanobis <- mahalanobis(
  wineDataVariables,
  center = colMeans(wineDataVariables),
  cov = cov(wineDataVariables)
)
wineData$surprise <- pchisq(wineData$mahalanobis, df = 23, lower.tail = FALSE)

wineData |>
  select(Sample, Cultivar, mahalanobis, surprise) |>
  mutate(
    mahalanobis = round(mahalanobis, 2),
    surprise = round(surprise, 4)
  ) |>
  rename(Wine = Sample, `Mahalanobis distance` = mahalanobis, `Tail probability` = surprise) |>
  arrange(desc(`Mahalanobis distance`)) |>
  knitr::kable()
```

```{r}
#| label: tbl-shapiro-pca
#| tbl-cap: "Shapiro-Wilk tests of normality applied to each principal component."

winePCA <- prcomp(wineDataVariables, center = TRUE, scale = TRUE)
wineDataPCADF <- as.data.frame(winePCA$x)

tibble(
  Component = colnames(wineDataPCADF),
  W = sapply(wineDataPCADF, \(x) shapiro.test(x)$statistic),
  `p-value` = sapply(wineDataPCADF, \(x) shapiro.test(x)$p.value)
) |>
  mutate(W = round(W, 3), `p-value` = round(`p-value`, 5)) |>
  knitr::kable()
```

```{r}
#| label: tbl-group-rank
#| tbl-cap: "Sample size, number of variables, and covariance matrix rank for each cultivar."

menciaData   <- wineData[wineData$Cultivar == "Mencia", 3:25]
alicanteData <- wineData[wineData$Cultivar == "Alicante Bouschet", 3:25]

tibble(
  Cultivar = c("Mencia", "Alicante Bouschet"),
  n = c(nrow(menciaData), nrow(alicanteData)),
  p = c(ncol(menciaData), ncol(alicanteData)),
  `Max rank (n-1)` = n - 1,
  `Covariance rank` = c(qr(cov(menciaData))$rank, qr(cov(alicanteData))$rank)
) |>
  knitr::kable()
```

Within-cultivar Mahalanobis distances cannot be computed on the full variable set. The covariance matrix of a group of $n$ observations has rank at most $n-1$, so with 20 Mencía and 10 Alicante Bouschet wines against 23 variables, both within-group covariance matrices are singular and cannot be inverted (@tbl-group-rank). We therefore restrict the within-group analysis to the leading principal components of each cultivar, retaining enough components that the covariance matrix is of full rank.

```{r}
#| label: fig-scree-groups
#| fig-cap: "Proportion of variance explained by each principal component, computed separately within each cultivar."
#| fig-width: 7
#| fig-height: 4

scree_data <- function(dat, cultivar) {
  p <- prcomp(dat, center = TRUE, scale. = TRUE)
  tibble(
    Cultivar = cultivar,
    PC = seq_along(p$sdev),
    Proportion = p$sdev^2 / sum(p$sdev^2)
  )
}

bind_rows(
  scree_data(menciaData, "Mencia"),
  scree_data(alicanteData, "Alicante Bouschet")
) |>
  filter(PC <= 10) |>
  ggplot(aes(PC, Proportion)) +
  geom_col(fill = "grey40") +
  facet_wrap(~Cultivar) +
  scale_x_continuous(breaks = 1:10) +
  labs(x = "Principal component", y = "Proportion of variance explained") +
  theme_minimal()
```

```{r}
#| label: tbl-mahalanobis-groupwise
#| tbl-cap: "Within-cultivar Mahalanobis distances computed on the leading principal components of each group."

pca_mahal <- function(dat, k) {
  scores <- prcomp(dat, center = TRUE, scale. = TRUE)$x[, 1:k, drop = FALSE]
  mahalanobis(scores, center = colMeans(scores), cov = cov(scores))
}

bind_rows(
  tibble(
    Wine = wineData$Sample[wineData$Cultivar == "Mencia"],
    Cultivar = "Mencia",
    Distance = pca_mahal(menciaData, 2)
  ),
  tibble(
    Wine = wineData$Sample[wineData$Cultivar == "Alicante Bouschet"],
    Cultivar = "Alicante Bouschet",
    Distance = pca_mahal(alicanteData, 3)
  )
) |>
  mutate(Distance = round(Distance, 2)) |>
  arrange(Cultivar, desc(Distance)) |>
  knitr::kable()
```

### Principal Component Analysis \[unfinished\]

As seen in @tbl-scree-R, the first five PCs each retained variance greater than 1, meaning they each account for more of the sample variance than the average of the original variables. The first five PCs also collectively retain 90.9% of the sample variance, while being a small enough number of variables to easily work with. As such, we retained the first five PCs and discarded the rest.

```{r}
#| label: tbl-scree-R
#| tbl-cap: "Variance and retained sample variance of PC1-PC6."

#summary table of PCs
library(pander)
wineDataVariables = as.data.frame(wine)[,3:25]
PCs = prcomp(wineDataVariables, center = TRUE, scale = TRUE)
summ = summary(PCs)
SDs = summ$importance[1,]
cumVar = summ$importance[3,]
PCtable = data.frame(PC = colnames(summ$importance)[1:6], SD = SDs[1:6], Cvar = cumVar[1:6])
rownames(PCtable) = NULL
colnames(PCtable)[3] = "Cumulative Variance"
pander(PCtable)
```

### Normality Assessment \[unfinished\] {#sec-normality-R}

Shapiro-Wilk tests on the derived PCs showed sufficient evidence to claim that PC1 (p=.0001) and PC5 (p=.01) were not normally distributed across the overall dataset. When assessed within each cultivar, both of these PCs appeared to follow a normal distribution, but PC2 did not appear to show a normal distribution within the Mencia (p=.0161) or Alicante Bouschet (p=.0095) cultivars (@tbl-shap-R) .

```{r}
#| label: tbl-shap-R
#| tbl-cap: "Shapiro-Wilk testing of PCs in aggregate and by cultivar."

# Shapiro-Wilk test of PC1-5 aggregate and by cultivar
library(pander)
aggShap = rep(NA,5)
menShap = rep(NA,5)
abShap = rep(NA,5)
for(i in 1:5){
  aggShap[i] = shapiro.test(wineDataPCADF[,i])$p.value
  menShap[i] = shapiro.test(wineDataPCADF[1:20,i])$p.value
  abShap[i] = shapiro.test(wineDataPCADF[21:30,i])$p.value
}
shapDF = data.frame(Component = c("PC1","PC2","PC3","PC4","PC5"), Aggregate = aggShap, Mencia = menShap, AlicanteBouschet = abShap)
colnames(shapDF)[4] = "Alicante Bouschet"

for(i in 1:5){
  for(j in 2:4){
    charTemp = as.character(round(as.numeric(shapDF[i,j]),4))
    shapDF[i,j] = ifelse(shapDF[i,j] <= 0.01, paste0(charTemp," ***"),
                        ifelse(shapDF[i,j] <= 0.05, paste0(charTemp, " **"),
                               ifelse(shapDF[i,j] <=0.1, paste0(charTemp, " *"),
                                      charTemp)))
  }
}
pander(shapDF)
```

### Graphical Assessment \[unfinished\]

PC1 and PC2 for the whole dataset are plotted on @fig-fullscatter-R . The separation by cultivar observed by [@revilla2016] is clearly replicated here. Samples M16 and M17 are noted as probable outliers within the Mencia cultivar, but there does not appear to be any indication of subgroupings within the Mencia samples. The Alicante Bouschet Samples, however, show a clear separation, with four samples diverging from the rest along PC2. Sample T22 also appears to be an outlier relative to the comparatively very tight clustering of T21 and T27-T30.

```{r}
#| label: fig-fullscatter-R
#| fig-cap: "Scatter plot of PC1 vs PC2 on full dataset."
#| results: hide

# graphical of PCs, full dataset
plot(wineDataPCADF$PC1,
     wineDataPCADF$PC2,
     col = c(rep("red",20),rep("blue",10)),
     lwd = 3,
     xlab = "PC1",
     ylab = "PC2",
     main = "Distribution of PC1 and PC2 Across Full Dataset") + 
  text(wineDataPCADF$PC1+0.1, wineDataPCADF$PC2-0.2, wineData$Sample, cex = 0.5)
```

### Surprising Observations Analysis \[unfinished\]

Under the assumption that all samples share an underlying multivariate normal distribution, sample M8 is a significant outlier (p=.0096), while samples M17 (p=.0747) and T23 (p=.0987) are slightly but not severely deviant from the assumed distribution (@tbl-mahalfull-R) . When each cultivar is assumed to form a separate multivariate normal distribution, however, there are no strong outliers and three slightly surprising samples: M7 (p=.0732), M8 (p=.0583), and M17(p=.0579). This lends moderate additional evidence to the claim that the variables measured by [@revilla2016] can reliably separate wine samples of these two cultivars. More compellingly, we see that every sample is highly inconsistent with the distribution of the opposing cultivar (p=.0057 for sample T25, p\<.0001 for all other samples), hence it is very unlikely that samples of either of these cultivars could be mistakenly categorised as belonging to the other cultivar when these measurements are known.

The final column of @tbl-mahalfull-R shows the p-value of the sample's Mahalanobis distance to the distribution of its cultivar, when the sample itself is excluded from calculating the distribution. Using this method, two of the Mencia samples appear slightly surprising: M16 (p=.0847) and M19 (p=.0811), one is surprising: M6 (p=.0367), and three are very surprising: M7 (p=.0002), M8 (p\<.0001) and M17 (p\<.0001). Within the Alicante Bouschet cultivar, however, samples T22-T26 all become very surprising (p=.0002 for T24, p\<.0001 for all others). This is consistent with these observations belonging to a separate distribution with greater variance, so the following analyses will focus on testing whether these five observations can sensibly be categorised as a separate cluster from samples T21 and T27-T30.

```{r}
#| label: tbl-mahalfull-R
#| tbl-cap: "Consistency of observations with theorised distributions."

# surprising results for mahalanobis: full model, cultivar split, cultivar split with exclusion
presFullModel = rep(NA,30)
presMenInc = rep(NA,30)
presABInc = rep(NA,30)
presSplitExc = rep(NA,30)
wineDataPCADFR = wineDataPCADF[,1:5]
PCAMahalanobis = mahalanobis(wineDataPCADFR,center = colMeans(wineDataPCADFR),cov = cov(wineDataPCADFR))
surprise = pchisq(PCAMahalanobis,df=5,lower.tail = FALSE)
PCAFullMenciaMahal = mahalanobis(wineDataPCADFR, center=colMeans(wineDataPCADFR[1:20,]), cov=cov(wineDataPCADFR[1:20,]))
surpriseMencia = pchisq(PCAFullMenciaMahal, df=5, lower.tail=FALSE)
PCAFullAlicanteMahal = mahalanobis(wineDataPCADFR, center=colMeans(wineDataPCADFR[21:30,]), cov=cov(wineDataPCADFR[21:30,]))
surpriseAlicante = pchisq(PCAFullAlicanteMahal, df=5, lower.tail=FALSE)
surpriseCulExc = rep(NA,30)
mahalCulExc = rep(NA,30)
for(i in 1:20){
  wineTemp = wineDataPCADFR[1:20,][-i,]
  muTemp = colMeans(wineTemp)
  covTemp = cov(wineTemp)
  mahalCulExc[i] = mahalanobis(wineDataPCADFR[i,], center = muTemp, cov=covTemp)
  surpriseCulExc[i] = pchisq(mahalCulExc[i], df=5, lower.tail=FALSE)
}
for(i in 21:30){
  iAdj = i - 20
  wineTemp = wineDataPCADFR[21:30,][-iAdj,]
  muTemp = colMeans(wineTemp)
  covTemp = cov(wineTemp)
  mahalCulExc[i] = mahalanobis(wineDataPCADFR[i,], center = muTemp, cov=covTemp)
  surpriseCulExc[i] = pchisq(mahalCulExc[i], df=5, lower.tail=FALSE)
}

for(i in 1:30){
  if(surprise[i] < 0.0001){
    presFullModel[i] = "< 0.0001 ***"
  } else if(surprise[i] <= 0.01){
    presFullModel[i] = paste0(round(surprise[i],4)," ***")
  } else if(surprise[i] <= 0.05){
    presFullModel[i] = paste0(round(surprise[i],4)," **")
  } else if(surprise[i] <= 0.1){
    presFullModel[i] = paste0(round(surprise[i],4)," *")
  } else {
    presFullModel[i] = as.character(round(surprise[i],4))
  }
  if(surpriseMencia[i] < 0.0001){
    presMenInc[i] = "< 0.0001 ***"
  } else if(surpriseMencia[i] <= 0.01){
    presMenInc[i] = paste0(round(surpriseMencia[i],4)," ***")
  } else if(surpriseMencia[i] <= 0.05){
    presMenInc[i] = paste0(round(surpriseMencia[i],4)," **")
  } else if(surpriseMencia[i] <= 0.1){
    presMenInc[i] = paste0(round(surpriseMencia[i],4)," *")
  } else {
    presMenInc[i] = as.character(round(surpriseMencia[i],4))
  }
  if(surpriseAlicante[i] < 0.0001){
    presABInc[i] = "< 0.0001 ***"
  } else if(surpriseAlicante[i] <= 0.01){
    presABInc[i] = paste0(round(surpriseAlicante[i],4)," ***")
  } else if(surpriseAlicante[i] <= 0.05){
    presABInc[i] = paste0(round(surpriseAlicante[i],4)," **")
  } else if(surpriseAlicante[i] <= 0.1){
    presABInc[i] = paste0(round(surpriseAlicante[i],4)," *")
  } else {
    presABInc[i] = as.character(round(surpriseAlicante[i],4))
  }
  if(surpriseCulExc[i] < 0.0001){
    presSplitExc[i] = "< 0.0001 ***"
  } else if(surpriseCulExc[i] <= 0.01){
    presSplitExc[i] = paste0(round(surpriseCulExc[i],4)," ***")
  } else if(surpriseCulExc[i] <= 0.05){
    presSplitExc[i] = paste0(round(surpriseCulExc[i],4)," **")
  } else if(surpriseCulExc[i] <= 0.1){
    presSplitExc[i] = paste0(round(surpriseCulExc[i],4)," *")
  } else {
    presSplitExc[i] = as.character(round(surpriseCulExc[i],4))
  }
}
mahalTable = data.frame(sample = wineData$Sample, fullModel = presFullModel, menciaInc = presMenInc, abInc = presABInc, splitExc = presSplitExc)
mahalTable$splitExc = gsub("2e-04","0.0002",mahalTable$splitExc)
colnames(mahalTable) = c("Sample","Full Dataset","Mencia","Alicante Bouschet","Cultivar Excluding Sample")
pander(mahalTable)
```

### Subgrouping Analysis \[unfinished\]

For the following analyses, samples T21 and T27-T30 are labelled as "Cluster 1" and samples T22-T26 are labelled "Cluster 2." PC4 and PC5 were dropped for these analyses, since the Mahalanobis distance cannot be meaningfully calculated for a cluster of five observations with more than three variables [@pires2024] . Collectively, these three PCs retain 79.7% of the sample variance {table PCs}.

Visual assessment of @fig-scatterAB-R indicates no overlap between Cluster 1 and Cluster 2, though Cluster 1 shows much tighter grouping than Cluster 2 and sample T22 has a PC2 value much more consistent with Cluster 1 than Cluster 2.

The Mahalanobis distances strongly indicate intra-cluster cohesion and inter-cluster separation, as all samples are highly consistent with the distribution of their theorised cluster (p\>=.2019 for all samples) and highly inconsistent with the distribution of the other theorised cluster (p\<.0001 for all samples) (@tbl-mahalAB-R). If the samples in Cluster 2 were simply random outliers of Cluster 1, we would expect the Cluster 1 samples to be consistent with the Cluster 2 distribution. Since this is not the case, we have evidence that these are two distinct categories of samples with separate distributions.

```{r}
#| label: fig-scatterAB-R
#| tbl-cap: "Scatter plot of PC1 vs PC2 within Alicante Bouschet samples."


# graph PC1-2 AB split
plot(wineDataPCADFR[21:30,]$PC1,
     wineDataPCADFR[21:30,]$PC2,
     col = c("green",rep("black",5),rep("green",4)),
     pch = as.character(c(1:9,0)))
```

```{r}
#| label: tbl-mahalAB-R
#| tbl-cap: "Consistency of Alicante Bouschet samples with theorised clusters."


# mahalanobis AB PC1-3 without exclusion
presCluster1 = rep(NA,10)
presCluster2 = rep(NA,10)

wineAB1 = wineDataPCADFR[c(21,27:30),1:3]
wineAB2 = wineDataPCADFR[22:26,1:3]
muAB1 = colMeans(wineAB1)
muAB2 = colMeans(wineAB2)
sigAB1 = cov(wineAB1)
sigAB2 = cov(wineAB2)

clus1from1surp = pchisq(mahalanobis(wineAB1,center = muAB1, cov = sigAB1), df=2, lower.tail = FALSE)
clus2from1surp = pchisq(mahalanobis(wineAB2,center = muAB1, cov = sigAB1), df=2, lower.tail = FALSE)
clus2from2surp = pchisq(mahalanobis(wineAB2,center = muAB2, cov = sigAB2), df=2, lower.tail = FALSE)
clus1from2surp = pchisq(mahalanobis(wineAB1,center = muAB2, cov = sigAB2), df=2, lower.tail = FALSE)

clus1Surp = c(clus1from1surp[1],clus2from1surp,clus1from1surp[2:5])
clus2Surp = c(clus1from2surp[1],clus2from2surp,clus1from2surp[2:5])



for(i in 1:10){
  if(clus1Surp[i] < 0.0001){
    presCluster1[i] = "< 0.0001 ***"
  } else if(clus1Surp[i] <= 0.01){
    presCluster1[i] = paste0(round(clus1Surp[i],4)," ***")
  } else if(clus1Surp[i] <= 0.05){
    presCluster1[i] = paste0(round(clus1Surp[i],4)," **")
  } else if(clus1Surp[i] <= 0.1){
    presCluster1[i] = paste0(round(clus1Surp[i],4)," *")
  } else {
    presCluster1[i] = as.character(round(clus1Surp[i],4))
  }
  if(clus2Surp[i] < 0.0001){
    presCluster2[i] = "< 0.0001 ***"
  } else if(clus2Surp[i] <= 0.01){
    presCluster2[i] = paste0(round(clus2Surp[i],4)," ***")
  } else if(clus2Surp[i] <= 0.05){
    presCluster2[i] = paste0(round(clus2Surp[i],4)," **")
  } else if(clus2Surp[i] <= 0.1){
    presCluster2[i] = paste0(round(clus2Surp[i],4)," *")
  } else {
    presCluster2[i] = as.character(round(clus2Surp[i],4))
  }
}
ABDF = data.frame(Sample = wineData$Sample[21:30], Cluster1 = presCluster1, Cluster2 = presCluster2)
pander(ABDF)
```

## Discussion {#sec-discussion}

The analysis set out to describe the phenolic profiles of 30 AOC Valdeorras red wines examined in [@revilla2016], assess whether a multivariate normal model appropriately describes the distribution of key chemical markers, identify atypical observations and determines whether the two cultivars separated cleanly on the 23-variable phenolic profile. The results broadly confirm the findings of Revilla *et al.* [@revilla2016] while surfacing a structural feature within the Alicante Bouschet group that their analysis did not address directly.

**Multivariate normality.** The correlation and univariate normality analyses (@sec-graphical, @sec-normality) show both that the raw phenolic variables are highly redundant, with 51 of 253 pairs correlating at \|r\| \> 0.8, and that several variables were non-normally distributed even on a within-cultivar level. 12 of the 42 within-cultivar Shapiro-Wilk tests rejected normality. These were concentrated within the Mencía cultivar (9 out of 12) and the coumaroylated anthocyanin derivatives of both cultivars. Since marginal normality is necessary for multivariate normality, we concluded neither cultivar’s phenolic profile is well approximated by a multivariate normal distribution on the raw variable scale. This, as well as the singular covariance matrices resulting from too many variables being assessed relative to the sample size, motivated the use of Principle Component Analysis throughout the remaining analysis. We were thereby able to work in a reduced space where the normality assumption is defensible. The pooled PCA scores (@tbl-shapiro-pca) were largely consistent with normality, with the exception of PC1 and PC5 overall and PC2 within each cultivar.

**Cultivar Separation.** Consistent with [@revilla2016], we found that a PC1-PC2 scatterplot of the overall dataset (@fig-fullscatter-R) shows a clear separation between Mencía and Alicante Bouschet samples. This is reinforced by the Mahalanobis distance analysis (@tbl-mahalfull-R), which showed that every sample is highly inconsistent with the distribution of the opposing cultivar (p \< 0.0001 for all but one sample), while within-cultivar Mahalanobis distances flag only a small number of mildly unusual wines (M7, M8 and M17). This reinforces that the phenolic measurements analysed here can reliably distinguish between the 2 cultivars.

**The sub-group structure within Alicante Bouschet.** The most notable finding to emerge beyond Revilla *et al.*'s original analysis [@revilla2016] is a clear internal split within the Alicante Bouschet samples. The PC1-PC2 scatterplot restricted to this cultivar (@fig-scatterAB-R) shows a tight clustering of samples T21 and T27-T30, while samples T22-T26 strongly diverge from this grouping. This separation is not a marginal effect: the Mahalanobis distance analysis (@tbl-mahalAB-R) shows every sample is highly consistent with its own theorised cluster (p ≥ 0.202) and highly inconsistent with the other cluster (p \< 0.0001 in every case). This is not consistent with samples T22-T26 being random outliers, since we would then expect the Cluster 1 samples to be largely consistent with the distribution of Cluster 2. The fact that the inconsistency is mutual lends credence to the claim that these are separate distributions.

This also helps us to explain a result flagged earlier in the report that the Alicante Bouschet samples show much greater variability on the concentration variables than the Mencía samples (@sec-descriptive), which [@revilla2016] attributed to differences in maceration length for wines intended for blending. The sub-group structure identified here offers a more specific account of that variability, where the Alicante Bouschet cultivar may fall into two discrete phenolic profiles. So plausibly Alicante Bouschet may have two different winemaking methods through maceration or distinct blends in different batches. As we have no record of which wines were produced under which conditions, this is speculation that we do not have a means to resolve here. We simply note this as an avenue of investigation, should one be interested in more precisely delineating subtypes of Alicante Bouschet wine for different blending goals.

**Limitations.** The Alicante Bouschet subgroup analysis should be read with caution given the very small sample count. The analysis found five wines per theorised sub-cluster, tested against covariance matrices estimated from those same five observations restricted to three principal components. With very few degrees of freedom, the Mahalanobis distance-based separation is based on a small enough sample that it should be treated as suggestive rather than confirmatory. Similarly, the non-normality of PC2 identified in @sec-normality-R means that the chi-square reference distribution underlying all Mahalanobis distance-based p-values in this report are only approximately valid. So the results should be interpreted as indicative of relative atypicality rather than exact inferential probabilities.

**Conclusion.** The combined 23-variable phenolic profiles reliably separate Mencía from Alicante Bouschet wines, just as in [@revilla2016]. The data suggests that Alicante Bouschet is not phenolically homogenous, with the samples splitting evenly into two distinct sub-clusters with a degree of separation comparable to the separation in the two cultivars. But given the overall small sample size for the data set as a whole (n = 30) and the cultivars' unbalanced distribution of samples (Mencía = 20, Alicante Bouschet = 10) this should be treated as a hypothesis-generating result rather than a confirmed statistical finding.
