---
title: "Early Indications of Subtyping within Alicante Bouschet Wines"
subtitle: "AOC Valdeorras, Galicia, Spain — STAT394 Group Report"
author:
  - "Riley Phillips"
  - "Wian de Vos"
  - "Oliver Temel"
  - "Simon Nelson"
format:
  revealjs:
    theme: serif
    slide-number: true
    transition: slide
    width: 1050
    height: 700
    toc: false
    css: |
      .title {
        font-size: 1em !important;
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

## Background {.smaller}

- Red wine colour, bitterness, astringency based on phenols, including anthocyanins
- Anthocyanins originate in grapes, produced based on genetics
- Ratio of different anthocyanins in wine ("anthocyanin fingerprint") strongly linked to grape cultivar

> Key takeaway: anthocyanin fingerprint of a wine should tell us what grape cultivar it was made  from

---

## The paper {.smaller}

- @revilla2016 compared wine from Mencia grapes (20 samples) with wine from Alicante Bouschet grapes (10 samples)
- UV-Vis Spectrophotometry (9 variables) compared to HPLC-DAD (14 variables) using PCA and LDA
- Both techniques separated samples by cultivar, HPLC more reliable

**Our aims**

- Look at grouping structure in data from @revilla2016
- Use aggregate data instead of separate blocks
- See if cultivar separation is replicated, and if any other structures are present

---

## Methods {.smaller}

- Descriptive analysis: compared mean and variance across cultivars
- Correlation analysis: Pearson correlations calculated for all pairs of raw variables
- Normality check: Shapiro-Wilk tests on all raw variables
- PCA: retained all principal components with variance greater than 1
- Mahalanobis distance: calculated the Mahalanobis distance for each point under four different conditions
- Cluster analysis: calculated Mahalanobis distance for each Alicante Bouschet sample assuming two clusters

:::notes
Riley finish 
:::

------------------------------------------------------------------------

## The data {.smaller}

- 30 wines: 20 Mencía, 10 Alicante Bouschet (2014 vintage, post-malolactic fermentation)
- 23 numeric variables combined from Revilla et al.'s Tables 1, 2, 5, 6:
  - 5 spectrophotometric/colour variables
  - 4 anthocyanin/tannin concentration variables (mg/L)
  - 14 relative-proportion anthocyanins (5 monoglucosides, 9 acylated derivatives)

------------------------------------------------------------------------

## A descriptive analysis {.smaller}

::::{.columns}
::: {.column width="60%"}
```{r}
library(dplyr)
library(tidyr)
#| label: tbl-summary-colour
#| tbl-cap: "Mean, standard deviation, and range of the spectrophotometric and colour variables by grape cultivar."

#| echo: false
#| eval: true
#| tbl-width: 7
#| tbl-height: 7

library(tidyverse)
library(ggcorrplot)

wine <- read_csv("extracted data.csv") |>
  select(-`...1`) |>
  mutate(
    Cultivar = factor(if_else(str_starts(Sample, "M"), "Mencia", "Alicante Bouschet")),
    .after = Sample
  )

wine |>
  select(Cultivar, Tot_Phen_Ind, Col_Int_420_520, Col_Int_420_520_620, Hue, Chem_Age_Ind) |>
  pivot_longer(-Cultivar, names_to = "Variable", values_to = "value") |>
  summarise(
    Mean = mean(value), SD = sd(value), Min = min(value), Max = max(value),
    .by = c(Cultivar, Variable)
  ) |>
  arrange(Variable, Cultivar) |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  knitr::kable()
```
:::
:::{.column width="40%"}

**Key descriptive pattern**: Alicante Bouschet consistently shows

- \~2× the colour intensity and total phenol index (teinturier character — pigmented flesh, not just skin)
- \~2–3× the anthocyanin/tannin concentrations, but with 2–3× the variability of Mencía
:::
::::
------------------------------------------------------------------------

## Redundancy among the 23 variables {.smaller}
```{r}
#| echo: false
#| eval: true
#| fig-width: 9
#| fig-height: 8

library(tidyverse)
library(ggcorrplot)

wine <- read_csv("extracted data.csv") |>
  select(-`...1`) |>
  mutate(
    Cultivar = factor(if_else(str_starts(Sample, "M"), "Mencia", "Alicante Bouschet")),
    .after = Sample
  )
```

::::{.columns}
::: {.column width="58%"}
```{r}
#| echo: false
#| label: fig-correlation
#| fig-cap: "Pairwise Pearson correlations among the 23 phenolic variables, ordered by hierarchical clustering."
#| fig-width: 6.5
#| fig-height: 5.5
wine |>
  select(-Sample, -Cultivar) |>
  cor() |>
  ggcorrplot(
    type = "lower",
    method = "circle",
    lab = FALSE,
    lab_size = 2,
    hc.order = TRUE,
    colors = c("blue", "white", "red"),
    legend.title = "Corr"
  )
```
:::
:::{.column width="42%"}
- **51 of 253 possible pairs (20.2%) correlate at \|r\| \> 0.8** — substantial redundancy
- Near-duplicate measurements: `Tot_Anthocyanins` vs `Tot_Anthocyans`, r = 0.998
- Compositional constraint: monoglucoside proportions sum to \~100%, so they're mechanically correlated (e.g. `MvGl`/`PnGl`, \|r\| ≈ 0.95–0.97)

> This redundancy motivates dimensionality reduction via PCA
:::

::: notes 

Oliver finish 
:::

::::
## Can we assume multivariate normality? {.smaller .scrollable}

:::::: columns
::: {.column width="42%"}
**Rejections at 5%** (13 of 46)

|     | Variable         | p      |
|-----|------------------|--------|
| M   | Tot_Anthocyanins | 0.0000 |
| M   | Tot_Anthocyans   | 0.0000 |
| M   | DpGlCm           | 0.0001 |
| M   | PtGlCm           | 0.0035 |
| M   | Tot_Tannins      | 0.0135 |
| AB  | DpGlCm           | 0.0159 |
| AB  | PnGlCm           | 0.0216 |
| M   | PnGlCm           | 0.0247 |
| M   | MvGlAc           | 0.0262 |
| M   | DpGl             | 0.0321 |
| M   | MvGlCf           | 0.0337 |
| AB  | PtGlAc           | 0.0427 |
| M   | PnGl             | 0.0482 |

*M = Mencía, AB = Alicante Bouschet*
:::

::: {.column width="6%"}
:::

::: {.column width="52%"}
**No — not on the raw variables.**

- **13 of 46** within-cultivar tests reject, against \~2 expected by chance
- **10 of 13** are Mencía
- Coumaroylated derivatives (DpGlCm, PnGlCm) reject in **both** cultivars
- PCA improves matters but **PC2 still rejects** in both

> Mahalanobis p-values that follow are approximately, not exactly, valid
:::
::::::

::: notes
This is the Shapiro-Wilk test — null hypothesis is that the data are normally distributed, so a small p-value means we reject normality. Chose it because n is small; it has better power than Kolmogorov-Smirnov at this sample size.

We tested each variable separately, within each cultivar. 23 variables × 2 cultivars = 46 tests.

The table shows all 13 rejections

10 of the 13 being Mencía sounds like Mencía is worse behaved, but Alicante Bouschet only has n=10 against Mencía's n=20, so we may just be less able to detect problems there.

What we did about it: rather than transform variables or abandon the normality-based approach, we moved to principal components. PCA gave us two things at once — it handled the redundancy, and the components are linear combinations of 23 variables, so the central limit theorem pushes them toward normality even when the inputs aren't normal. That's why PC1, PC3, PC4 all pass.

It didn't fully work — PC2 still rejects in both cultivars — so we proceeded with Mahalanobis distances knowing the chi-square reference distribution is approximate. We flag this in Limitations rather than pretending it away.

Wian finish
:::
------------------------------------------------------------------------

## Why PCA {.smaller}

:::: {.columns}

::: {.column width="48%"}
```{r}
#| echo: false
#| label: fig-scree
#| fig-cap: "Scree plot of the first ten principal components."
#| fig-width: 5.5
#| fig-height: 4

winePCA <- prcomp(wine |> select(-Sample, -Cultivar), center = TRUE, scale. = TRUE)

tibble(PC = 1:10, Eigenvalue = winePCA$sdev[1:10]^2) |>
  ggplot(aes(PC, Eigenvalue)) +
  geom_line(colour = "#2471A3", linewidth = 1) +
  geom_point(colour = "#2471A3", size = 3) +
  geom_hline(yintercept = 1, colour = "#C0392B", linetype = "dashed", linewidth = 0.7) +
  scale_x_continuous(breaks = 1:10) +
  labs(x = "Principal Component", y = "Eigenvalue") +
  theme_minimal(base_size = 13)
```
:::

::: {.column width="4%"}
:::

::: {.column width="48%"}
Two independent problems point to the same fix:

1. **Singular covariance matrices** — 23 variables, only 20 (or 10) wines per cultivar, so $\hat\Sigma$ can't be inverted and Mahalanobis distance is undefined
2. **Redundancy** — much of the 23-dimensional variation is not independent
:::

::::

**Approach**: retain PCs with eigenvalue > 1 (Kaiser criterion) — the first 5 retain **90.9%** of total variance, giving an invertible space for Mahalanobis distances and normality checks

::: {.notes}
Eigenvalue > 1 means a component accounts for more variance than a single original
variable would on its own — that's the Kaiser criterion, and the dashed line is where
it sits.

Be honest about the cutoff: PC1 is dominant, PC2 and PC3 are clearly above the line,
but PC4 and PC5 are only just above it. There's no dramatic elbow here — we followed
the rule rather than reading an obvious break. If someone pushes, the answer is that
dropping to three PCs would still leave us able to invert the covariance matrix, but
we'd lose about 11% of the variance and PC4 turns out to matter for M7.

Both problems on this slide are really the same problem in different clothes —
23 variables is too many for 30 observations, and they're not independent anyway.

Next slide checks whether the components are better behaved than the raw variables.

Oliver finish 
:::

------------------------------------------------------------------------

## Normality of the principal components {.smaller}

```{r}
#| echo: false
#| eval: true


library(pander)
aggShap = rep(NA,5)
menShap = rep(NA,5)
abShap = rep(NA,5)
wineData = as.data.frame(wine)
wineDataVariables = wineData[, 3:25]
winePCA <- prcomp(wineDataVariables, center = TRUE, scale = TRUE)
wineDataPCADF <- as.data.frame(winePCA$x)

for(i in 1:5){
  aggShap[i] = shapiro.test(wineDataPCADF[,i])$p.value
  menShap[i] = shapiro.test(wineDataPCADF[1:20,i])$p.value
  abShap[i] = shapiro.test(wineDataPCADF[21:30,i])$p.value
}
shapDF = data.frame(Component = c("PC1","PC2","PC3","PC4","PC5"), Aggregate = aggShap, Mencia = menShap, AlicanteBouschet = abShap)
colnames(shapDF)[4] = "Alicante Bouschet"

for(i in 1:5){
  for(j in 2:4){
    charTemp = as.character(round(as.numeric(shapDF[i,j]),4))
    shapDF[i,j] = ifelse(shapDF[i,j] <= 0.01, paste0(charTemp," ***"),
                        ifelse(shapDF[i,j] <= 0.05, paste0(charTemp, " **"),
                               ifelse(shapDF[i,j] <=0.1, paste0(charTemp, " *"),
                                      charTemp)))
  }
}
shapDF[1,2] = "0.0001 ***"

pander::pander(shapDF)  # PC1–5 Shapiro-Wilk, aggregate + by cultivar
```

- Pooled: **PC1** (p = 0.0001) and **PC5** (p = 0.01) reject normality
- Within cultivar: PC1 and PC5 are fine, **but PC2 rejects in both** cultivars (Mencía p = 0.0161, Alicante Bouschet p = 0.0095)

> Reducing dimensionality helped, but did **not** fully resolve non-normality — Mahalanobis-based p-values later on should be read as approximately, not exactly, valid

------------------------------------------------------------------------

## Can we find groups? {.smaller}

::::: columns
::: {.column width="55%"}
```{r}
#| echo: false
#| label: fig-pca-cultivar
#| fig-cap: "Wines projected onto the first two principal components, coloured by cultivar."
#| fig-width: 6
#| fig-height: 5

winePCA <- prcomp(wine |> select(-Sample, -Cultivar), center = TRUE, scale. = TRUE)

as_tibble(winePCA$x[, 1:2]) |>
  mutate(Cultivar = wine$Cultivar, Wine = wine$Sample) |>
  ggplot(aes(PC1, PC2, colour = Cultivar)) +
  geom_point(size = 3) +
  geom_text(aes(label = Wine), vjust = -1, size = 3, show.legend = FALSE) +
  scale_colour_manual(values = c("Mencia" = "#C0392B", "Alicante Bouschet" = "#2471A3")) +
  labs(x = "PC1", y = "PC2") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")
```
:::

::: {.column width="45%"}
**Yes — and more than we expected.**

- Mencía / Alicante Bouschet split recovered cleanly on PC1
- A **second split** appears within Alicante Bouschet along PC2
- Mahalanobis evidence is mutual (p \< 0.0001 both directions)
:::
:::::

::: notes
Left cluster is Mencía, right is Alicante Bouschet — that's PC1 doing the work, and it's the separation Revilla et al. already found.

The new bit is vertical. Within the blue points, T23/T24/T25/T26 sit high on PC2 while T21/T27/T28/T29/T30 sit low. T22 is the awkward one — it's low on PC2 like Cluster 1 but far left on PC1, so it doesn't sit neatly in either.

if T22-T26 were just outliers, Cluster 1 wines would still look consistent with Cluster 2's distribution. They don't — p \< 0.0001 in both directions. That's what distinguishes a subgroup from a tail.

M16 and M17 bottom left are the Mencía outliers from the descriptive section — the two with high anthocyanins and low tannins.

:::
------------------------------------------------------------------------

## Are the groups consistent with graph type {.smaller}

**Partly.**

- The **primary** split is exactly grape type — every wine is highly inconsistent with the opposing cultivar's distribution (p \< 0.0001, one exception at p = 0.0057)
- Practically: a wine's phenolic profile alone would be enough to identify its cultivar
- But the **secondary** split cuts *within* Alicante Bouschet — grape type alone doesn't account for all the structure in the data

> Something other than cultiar is orginising these wines

::: notes
(read first bullet point) - this is exactly what Revilla et al got, using all 23 variables jointly rather than block by block. The one exception at p = 0.0057 is T25, which is still well below any threshold

this is why anthocyanin fingerprinting is used for authentication in the first place, so it connects back to the motivation slide.

the point is that if grape type were the whole story, Alicante Bouschet would look like a single tight group. It doesn't.
:::

---

## Surprising observations: individual wines {.smaller}

:::: {.columns}

::: {.column width="52%"}
```{r}
#| echo: false
#| label: fig-mencia-pc45
#| fig-cap: "Mencía wines on the fourth and fifth principal components, with M7 and M8 highlighted."
#| fig-width: 6
#| fig-height: 5

as_tibble(winePCA$x[wine$Cultivar == "Mencia", 4:5]) |>
  mutate(
    Wine = wine$Sample[wine$Cultivar == "Mencia"],
    Flagged = Wine %in% c("M7", "M8")
  ) |>
  ggplot(aes(PC4, PC5, colour = Flagged)) +
  geom_point(size = 3) +
  geom_text(aes(label = Wine), vjust = -1, size = 3, show.legend = FALSE) +
  scale_colour_manual(values = c("TRUE" = "#C0392B", "FALSE" = "grey60")) +
  labs(x = "PC4", y = "PC5") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```
:::

::: {.column width="6%"}
:::

::: {.column width="42%"}
**Two Mencía wines flagged: M7, M8**

- M8 is the strongest outlier on the pooled analysis (p = 0.0096)
- None of the two stand out on PC1–PC2 — the view we've been looking at
- M7 sits at **+3.18 on PC4**; M8 at **−3.52 on PC5**

A two-dimensional plot shows two of the five directions we retained — Mahalanobis distance uses all of them
:::

::::

::: {.notes}
Start by admitting the obvious: if you went back to the earlier PC1-PC2 plot, M7 and
M8 sit right in the middle of the Mencía cloud. They look completely ordinary.

That's exactly why we didn't judge by eye. PC4 and PC5 have standard deviations of
about 1.2 and 1.0, so +3.18 and -3.52 are roughly 2.6 and 3.4 SDs out — clearly
extreme, just in directions the default view doesn't show.

M17 is the one you could see earlier, bottom left with M16 — it's spread across PC2
and PC3.

The general point: Mahalanobis distance is doing work here that visual inspection
can't. Worth saying explicitly since the rest of the deck leans on scatterplots.

If asked why not just plot more component pairs — with five PCs that's ten plots, and
the distance measure summarises all of it in one number.
:::

---

##  A whole subgroup {.smaller}

:::: {.columns}

::: {.column width="52%"}
```{r}
#| echo: false
#| label: fig-ab-clusters
#| fig-cap: "Alicante Bouschet wines on the first two principal components, coloured by theorised cluster."
#| fig-width: 6
#| fig-height: 5

as_tibble(winePCA$x[wine$Cultivar == "Alicante Bouschet", 1:2]) |>
  mutate(
    Wine = wine$Sample[wine$Cultivar == "Alicante Bouschet"],
    Cluster = if_else(Wine %in% c("T22","T23","T24","T25","T26"), "Cluster 2", "Cluster 1")
  ) |>
  ggplot(aes(PC1, PC2, colour = Cluster)) +
  geom_point(size = 3) +
  geom_text(aes(label = Wine), vjust = -1, size = 3, show.legend = FALSE) +
  scale_colour_manual(values = c("Cluster 1" = "#27AE60", "Cluster 2" = "#2C3E50")) +
  labs(x = "PC1", y = "PC2") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```
:::

::: {.column width="6%"}
:::

::: {.column width="42%"}
**T22–T26 became *very* surprising — all five at once**

- Tested against their own cultivar **excluding themselves**, all five reject (p ≤ 0.0002)
- Five simultaneous outliers is not a plausible tail
- Treating them as a cluster: each wine consistent with its own (p ≥ 0.202), inconsistent with the other (p < 0.0001)

> What looks like a set of outliers under one assumption is a coherent subgroup under another
:::

::::

::: {.notes}
The exclusion method is what revealed this. Testing each wine against the full cultivar
distribution hid the effect, because those five were inflating the very variance they
were being compared against. Drop each one from the calculation and they light up.

Same data, same distance measure, different reference distribution, opposite conclusion.
That's the methodological point worth dwelling on.

Mutual inconsistency is what distinguishes a subgroup from a tail — if these were just
outliers, Cluster 1 wines would still look fine against Cluster 2's distribution. They
don't.

T22 is awkward: low on PC2 like Cluster 1 but far left on PC1. The three-PC test puts
it in Cluster 2; PC2 alone wouldn't. Be upfront if asked.

Don't oversell — we found this visually then tested it, which is partly circular. Save
for Limitations unless someone raises it.

Wian finish 
:::

------------------------------------------------------------------------

## Why might this be happening? {.smaller}

- Alicante Bouschet already showed 2–3× the variability of Mencía on concentration variables (descriptive stage)
- Revilla et al. attributed this to variable maceration length — the cultivar is often managed to different blending targets
- **Our finding refines this**: rather than one heterogeneous group with a long tail, the data are more consistent with **two discrete phenolic profiles**

> Plausible explanation: different maceration regimes or blending targets across batches — but we have no winemaking records to confirm this directly

------------------------------------------------------------------------

## Limitations {.smaller}

- Subgroup analysis rests on **5 wines per cluster**, tested against covariance estimated from those same 5 observations (3 PCs) — small enough to be suggestive, not confirmatory
- PC2's non-normality means the chi-square reference distribution underlying all Mahalanobis p-values is only **approximately** valid
- Overall dataset is small (n=30) and unbalanced (20 vs. 10) — limits power throughout, especially for Alicante Bouschet-specific claims

:::notes 
Oliver finish 
:::
------------------------------------------------------------------------

## Conclusion: the four questions {.smaller}

Our analysis set out to answer four questions about these 30 wines:

1.  **Can we find groups in the data?**
2.  **Are they consistent with grape type?**
3.  **Can we assume multivariate normality?**
4.  **Are there surprising observations?**

:::notes 
One: yes, and more than we expected. The cultivar split came through cleanly on PC1,
and a second split appeared within Alicante Bouschet that Revilla et al. didn't
report.

Two: partly. The primary split is grape type — every wine rejects the opposing
cultivar's distribution. But the secondary split cuts within a cultivar, so grape
type alone doesn't explain all the structure.

Three: no, not on the raw variables — thirteen of forty-six within-cultivar tests
reject. PCA improved things but PC2 still rejects, so our Mahalanobis p-values are
approximate rather than exact.

Four: yes, at two levels. M7 and M8 individually, extreme on components a scatterplot
doesn't show. And T22 to T26 collectively, which turned out to be a subgroup rather
than a set of outliers.

Close on the honest framing: small sample, unbalanced, and we identified the clusters
visually before testing them. Hypothesis-generating, not confirmed. Worth testing in
a larger sample with maceration records.

Wian finish 

:::

---

## References {.smaller}
