---
title: "Early Indications of Subtyping within Alicante Bouschet Wines"
subtitle: "AOC Valdeorras, Galicia, Spain — STAT394 Group Report"
author:
  - "Riley Phillips"
  - "Wian de Vos"
  - "Oliver Temel"
  - "Simon Nelson"
format:
  revealjs:
    theme: serif
    slide-number: true
    transition: slide
    width: 1050
    height: 700
    toc: false
    css: |
      .title {
        font-size: 1em !important;
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

## Background {.smaller}

- Red wine colour, bitterness, astringency based on phenols, including anthocyanins
- Anthocyanins originate in grapes, produced based on genetics
- Ratio of different anthocyanins in wine ("anthocyanin fingerprint") strongly linked to grape cultivar

> Key takeaway: anthocyanin fingerprint of a wine should tell us what grape cultivar it was made  from

---

## The paper {.smaller}

- @revilla2016 compared wine from Mencia grapes (20 samples) with wine from Alicante Bouschet grapes (10 samples)
- UV-Vis Spectrophotometry (9 variables) compared to HPLC-DAD (14 variables) using PCA and LDA
- Both techniques separated samples by cultivar, HPLC more reliable

**Our aims**

- Look at grouping structure in data from @revilla2016
- Use aggregate data instead of separate blocks
- See if cultivar separation is replicated, and if any other structures are present

---

## Methods {.smaller}

- Descriptive analysis: compared mean and variance across cultivars
- Correlation analysis: Pearson correlations calculated for all pairs of raw variables
- Normality check: Shapiro-Wilk tests on all raw variables
- PCA: retained all principal components with variance greater than 1
- Mahalanobis distance: calculated the Mahalanobis distance for each point under four different conditions
- Cluster analysis: calculated Mahalanobis distance for each Alicante Bouschet sample assuming two clusters

:::notes
Riley finish 
:::

------------------------------------------------------------------------

## The data {.smaller}

- 30 wines: 20 Mencía, 10 Alicante Bouschet (2014 vintage, post-malolactic fermentation)
- 23 numeric variables combined from Revilla et al.'s Tables 1, 2, 5, 6:
  - 5 spectrophotometric/colour variables
  - 4 anthocyanin/tannin concentration variables (mg/L)
  - 14 relative-proportion anthocyanins (5 monoglucosides, 9 acylated derivatives)

------------------------------------------------------------------------

## A descriptive analysis {.smaller}

::::{.columns}
::: {.column width="60%"}
```{r}
library(dplyr)
library(tidyr)
#| label: tbl-summary-colour
#| tbl-cap: "Mean, standard deviation, and range of the spectrophotometric and colour variables by grape cultivar."

#| echo: false
#| eval: true
#| tbl-width: 7
#| tbl-height: 7

library(tidyverse)
library(ggcorrplot)

wine <- read_csv("extracted data.csv") |>
  select(-`...1`) |>
  mutate(
    Cultivar = factor(if_else(str_starts(Sample, "M"), "Mencia", "Alicante Bouschet")),
    .after = Sample
  )

wine |>
  select(Cultivar, Tot_Phen_Ind, Col_Int_420_520, Col_Int_420_520_620, Hue, Chem_Age_Ind) |>
  pivot_longer(-Cultivar, names_to = "Variable", values_to = "value") |>
  summarise(
    Mean = mean(value), SD = sd(value), Min = min(value), Max = max(value),
    .by = c(Cultivar, Variable)
  ) |>
  arrange(Variable, Cultivar) |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  knitr::kable()
```
:::
:::{.column width="40%"}

**Key descriptive pattern**: Alicante Bouschet consistently shows

- \~2× the colour intensity and total phenol index (teinturier character — pigmented flesh, not just skin)
- \~2–3× the anthocyanin/tannin concentrations, but with 2–3× the variability of Mencía
:::
::::
------------------------------------------------------------------------

## Redundancy among the 23 variables {.smaller}
```{r}
#| echo: false
#| eval: true
#| fig-width: 9
#| fig-height: 8

library(tidyverse)
library(ggcorrplot)

wine <- read_csv("extracted data.csv") |>
  select(-`...1`) |>
  mutate(
    Cultivar = factor(if_else(str_starts(Sample, "M"), "Mencia", "Alicante Bouschet")),
    .after = Sample
  )
```

::::{.columns}
::: {.column width="58%"}
```{r}
#| echo: false
#| label: fig-correlation
#| fig-cap: "Pairwise Pearson correlations among the 23 phenolic variables, ordered by hierarchical clustering."
#| fig-width: 6.5
#| fig-height: 5.5
wine |>
  select(-Sample, -Cultivar) |>
  cor() |>
  ggcorrplot(
    type = "lower",
    method = "circle",
    lab = FALSE,
    lab_size = 2,
    hc.order = TRUE,
    colors = c("blue", "white", "red"),
    legend.title = "Corr"
  )
```
:::
:::{.column width="42%"}
- **51 of 253 possible pairs (20.2%) correlate at \|r\| \> 0.8** — substantial redundancy
- Near-duplicate measurements: `Tot_Anthocyanins` vs `Tot_Anthocyans`, r = 0.998
- Compositional constraint: monoglucoside proportions sum to \~100%, so they're mechanically correlated (e.g. `MvGl`/`PnGl`, \|r\| ≈ 0.95–0.97)

> This redundancy motivates dimensionality reduction via PCA
:::

::: notes 

Oliver finish 
:::

::::
## Can we assume multivariate normality? {.smaller .scrollable}

:::::: columns
::: {.column width="42%"}
**Rejections at 5%** (13 of 46)

|     | Variable         | p      |
|-----|------------------|--------|
| M   | Tot_Anthocyanins | 0.0000 |
| M   | Tot_Anthocyans   | 0.0000 |
| M   | DpGlCm           | 0.0001 |
| M   | PtGlCm           | 0.0035 |
| M   | Tot_Tannins      | 0.0135 |
| AB  | DpGlCm           | 0.0159 |
| AB  | PnGlCm           | 0.0216 |
| M   | PnGlCm           | 0.0247 |
| M   | MvGlAc           | 0.0262 |
| M   | DpGl             | 0.0321 |
| M   | MvGlCf           | 0.0337 |
| AB  | PtGlAc           | 0.0427 |
| M   | PnGl             | 0.0482 |

*M = Mencía, AB = Alicante Bouschet*
:::

::: {.column width="6%"}
:::

::: {.column width="52%"}
**No — not on the raw variables.**

- **13 of 46** within-cultivar tests reject, against \~2 expected by chance
- **10 of 13** are Mencía
- Coumaroylated derivatives (DpGlCm, PnGlCm) reject in **both** cultivars
- PCA improves matters but **PC2 still rejects** in both

> Mahalanobis p-values that follow are approximately, not exactly, valid
:::
::::::

::: notes
This is the Shapiro-Wilk test — null hypothesis is that the data are normally distributed, so a small p-value means we reject normality. Chose it because n is small; it has better power than Kolmogorov-Smirnov at this sample size.

We tested each variable separately, within each cultivar. 23 variables × 2 cultivars = 46 tests.

The table shows all 13 rejections

10 of the 13 being Mencía sounds like Mencía is worse behaved, but Alicante Bouschet only has n=10 against Mencía's n=20, so we may just be less able to detect problems there.

Wian finish
:::
------------------------------------------------------------------------

## Why PCA {.smaller}

:::: {.columns}

::: {.column width="48%"}
```{r}
#| echo: false
#| label: fig-scree
#| fig-cap: "Scree plot of the first ten principal components."
#| fig-width: 5.5
#| fig-height: 4

winePCA <- prcomp(wine |> select(-Sample, -Cultivar), center = TRUE, scale. = TRUE)

tibble(PC = 1:10, Eigenvalue = winePCA$sdev[1:10]^2) |>
  ggplot(aes(PC, Eigenvalue)) +
  geom_line(colour = "#2471A3", linewidth = 1) +
  geom_point(colour = "#2471A3", size = 3) +
  geom_hline(yintercept = 1, colour = "#C0392B", linetype = "dashed", linewidth = 0.7) +
  scale_x_continuous(breaks = 1:10) +
  labs(x = "Principal Component", y = "Eigenvalue") +
  theme_minimal(base_size = 13)
```
:::

::: {.column width="4%"}
:::

::: {.column width="48%"}
Two independent problems point to the same fix:

1. **Singular covariance matrices** — 23 variables, only 20 (or 10) wines per cultivar, so $\hat\Sigma$ can't be inverted and Mahalanobis distance is undefined
2. **Redundancy** — much of the 23-dimensional variation is not independent
:::

::::

**Approach**: retain PCs with eigenvalue > 1 (Kaiser criterion) — the first 5 retain **90.9%** of total variance, giving an invertible space for Mahalanobis distances and normality checks

::: {.notes}


:::

------------------------------------------------------------------------

## Normality of the principal components {.smaller}

```{r}
#| echo: false
#| eval: true


library(pander)
aggShap = rep(NA,5)
menShap = rep(NA,5)
abShap = rep(NA,5)
wineData = as.data.frame(wine)
wineDataVariables = wineData[, 3:25]
winePCA <- prcomp(wineDataVariables, center = TRUE, scale = TRUE)
wineDataPCADF <- as.data.frame(winePCA$x)

for(i in 1:5){
  aggShap[i] = shapiro.test(wineDataPCADF[,i])$p.value
  menShap[i] = shapiro.test(wineDataPCADF[1:20,i])$p.value
  abShap[i] = shapiro.test(wineDataPCADF[21:30,i])$p.value
}
shapDF = data.frame(Component = c("PC1","PC2","PC3","PC4","PC5"), Aggregate = aggShap, Mencia = menShap, AlicanteBouschet = abShap)
colnames(shapDF)[4] = "Alicante Bouschet"

for(i in 1:5){
  for(j in 2:4){
    charTemp = as.character(round(as.numeric(shapDF[i,j]),4))
    shapDF[i,j] = ifelse(shapDF[i,j] <= 0.01, paste0(charTemp," ***"),
                        ifelse(shapDF[i,j] <= 0.05, paste0(charTemp, " **"),
                               ifelse(shapDF[i,j] <=0.1, paste0(charTemp, " *"),
                                      charTemp)))
  }
}
shapDF[1,2] = "0.0001 ***"

pander::pander(shapDF)  # PC1–5 Shapiro-Wilk, aggregate + by cultivar
```

- Pooled: **PC1** (p = 0.0001) and **PC5** (p = 0.01) reject normality
- Within cultivar: PC1 and PC5 are fine, **but PC2 rejects in both** cultivars (Mencía p = 0.0161, Alicante Bouschet p = 0.0095)

> Reducing dimensionality helped, but did **not** fully resolve non-normality — Mahalanobis-based p-values later on should be read as approximately, not exactly, valid

:::notes

Oliver finish

:::
------------------------------------------------------------------------

## Can we find groups? {.smaller}

::::: columns
::: {.column width="55%"}
```{r}
#| echo: false
#| label: fig-pca-cultivar
#| fig-cap: "Wines projected onto the first two principal components, coloured by cultivar."
#| fig-width: 6
#| fig-height: 5

winePCA <- prcomp(wine |> select(-Sample, -Cultivar), center = TRUE, scale. = TRUE)

as_tibble(winePCA$x[, 1:2]) |>
  mutate(Cultivar = wine$Cultivar, Wine = wine$Sample) |>
  ggplot(aes(PC1, PC2, colour = Cultivar)) +
  geom_point(size = 3) +
  geom_text(aes(label = Wine), vjust = -1, size = 3, show.legend = FALSE) +
  scale_colour_manual(values = c("Mencia" = "#C0392B", "Alicante Bouschet" = "#2471A3")) +
  labs(x = "PC1", y = "PC2") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")
```
:::

::: {.column width="45%"}
**Yes — and more than we expected.**

- Mencía / Alicante Bouschet split recovered cleanly on PC1
- A **second split** appears within Alicante Bouschet along PC2
:::
:::::

::: notes
Left cluster is Mencía, right is Alicante Bouschet — that's PC1 doing the work, and it's the separation Revilla et al. already found.

The new bit is vertical. Within the blue points, T23/T24/T25/T26 sit high on PC2 while T21/T27/T28/T29/T30 sit low. T22 is the awkward one — it's low on PC2 like Cluster 1 but far left on PC1, so it doesn't sit neatly in either.

:::
------------------------------------------------------------------------

## Are the groups consistent with grape type {.smaller}

**Partly.**

- The **primary** split is exactly grape type — every wine is highly inconsistent with the opposing cultivar's distribution (p \< 0.0001, one exception at p = 0.0057)
- Practically: a wine's phenolic profile alone would be enough to identify its cultivar
- But the **secondary** split cuts *within* Alicante Bouschet — grape type alone doesn't account for all the structure in the data

> Something other than cultiar is orginising these wines

::: notes
(read first bullet point) - this is exactly what Revilla et al got, using all 23 variables jointly rather than block by block. The one exception at p = 0.0057 is T25, which is still well below any threshold

this is why anthocyanin fingerprinting is used for authentication in the first place

the point is that if grape type were the whole story, Alicante Bouschet would look like a single tight group. It doesn't.
:::

---

## Surprising observations: individual wines {.smaller}

:::: {.columns}

::: {.column width="52%"}
```{r}
#| echo: false
#| label: fig-mencia-pc45
#| fig-cap: "Mencía wines on the fourth and fifth principal components, with M7 and M8 highlighted."
#| fig-width: 6
#| fig-height: 5

as_tibble(winePCA$x[wine$Cultivar == "Mencia", 4:5]) |>
  mutate(
    Wine = wine$Sample[wine$Cultivar == "Mencia"],
    Flagged = Wine %in% c("M7", "M8")
  ) |>
  ggplot(aes(PC4, PC5, colour = Flagged)) +
  geom_point(size = 3) +
  geom_text(aes(label = Wine), vjust = -1, size = 3, show.legend = FALSE) +
  scale_colour_manual(values = c("TRUE" = "#C0392B", "FALSE" = "grey60")) +
  labs(x = "PC4", y = "PC5") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```
:::

::: {.column width="6%"}
:::

::: {.column width="42%"}
**Two Mencía wines flagged: M7, M8**

- Neither stands out on PC1–PC2 — the view we've been using
- M7 sits at **+3.18 on PC4**, M8 at **−3.52 on PC5**
- M8 is the strongest outlier on the pooled analysis (p = 0.0096)

A two-dimensional plot shows two of the five directions we retained — Mahalanobis distance uses all of them
:::

::::

::: {.notes}
Start by admitting the obvious: if you went back to the earlier PC1-PC2 plot, M7 and
M8 sit right in the middle of the Mencía cloud. They look completely ordinary.

That's exactly why we didn't judge by eye. PC4 and PC5 have standard deviations of
about 1.2 and 1.0, so +3.18 and -3.52 are roughly 2.6 and 3.4 SDs out — clearly
extreme, just in directions the default view doesn't show.

M16 and M17 were the two visibly separated wines on the earlier PC1–PC2 plot — M17 also registers as mildly surprising; M7 and M8 are invisible there and only surface on PC4 and PC5

The general point: Mahalanobis distance is doing work here that visual inspection can't. It measures distance from the centre of the distribution, scaled by how spread out the data are in each direction — so a point can look ordinary on the dominant components and still be extreme once all five are accounted for

If asked why not just plot more component pairs — with five PCs that's ten plots, and
the distance measure summarises all of it in one number.
:::

---

##  A whole subgroup {.smaller}

:::: {.columns}

::: {.column width="52%"}
```{r}
#| echo: false
#| label: fig-ab-clusters
#| fig-cap: "Alicante Bouschet wines on the first two principal components, coloured by theorised cluster."
#| fig-width: 6
#| fig-height: 5

as_tibble(winePCA$x[wine$Cultivar == "Alicante Bouschet", 1:2]) |>
  mutate(
    Wine = wine$Sample[wine$Cultivar == "Alicante Bouschet"],
    Cluster = if_else(Wine %in% c("T22","T23","T24","T25","T26"), "Cluster 2", "Cluster 1")
  ) |>
  ggplot(aes(PC1, PC2, colour = Cluster)) +
  geom_point(size = 3) +
  geom_text(aes(label = Wine), vjust = -1, size = 3, show.legend = FALSE) +
  scale_colour_manual(values = c("Cluster 1" = "#27AE60", "Cluster 2" = "#2C3E50")) +
  labs(x = "PC1", y = "PC2") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```
:::

::: {.column width="6%"}
:::

::: {.column width="42%"}
**T22–T26 became *very* surprising — all five at once**

- Tested against their own cultivar **excluding themselves**, all five reject (p ≤ 0.0002)
- Five simultaneous outliers is not a plausible tail
- As clusters: each wine fits its own (p ≥ 0.202), rejects the other (p < 0.0001)

> Outliers under one assumption; a coherent subgroup under another
:::

::::

::: {.notes}
First we asked the ordinary question: is any single Alicante Bouschet wine unusual for
its cultivar? Nothing came up.

But including the wine in its own comparison is flawed — the wine being checked is one of the ten used to define "normal" in the first place. It pulls the mean toward itself and widens the spread, so it's compared against a standard it helped set. With ten wines, each is a tenth of that standard: enough to hide itself.

So we left each wine out of its own comparison. Five failed at once — T22 to T26, all
at p ≤ 0.0002. One failure is an outlier; five together isn't a tail. Collectively
they'd been stretching the spread wide enough to cover everyone.

So we tested the ten as two groups. Each wine fits its own cluster (p ≥ 0.202) and
rejects the other (p < 0.0001), and that works both directions — if these were just
stragglers, Cluster 1 would still look comfortable against Cluster 2's distribution.
It doesn't. Two groups, not one with a long tail.


Wian finish 
:::

------------------------------------------------------------------------

## Why might this be happening? {.smaller}

- Alicante Bouschet already showed 2–3× the variability of Mencía on concentration variables (descriptive stage)
- Revilla et al. attributed this to variable maceration length — the cultivar is often managed to different blending targets
- **Our finding refines this**: rather than one heterogeneous group with a long tail, the data are more consistent with **two discrete phenolic profiles**

> Plausible explanation: different maceration regimes or blending targets across batches — but we have no winemaking records to confirm this directly

------------------------------------------------------------------------

## Limitations {.smaller}

- Subgroup analysis rests on **5 wines per cluster**, tested against covariance estimated from those same 5 observations (3 PCs) — small enough to be suggestive, not confirmatory
- PC2's non-normality means the chi-square reference distribution underlying all Mahalanobis p-values is only **approximately** valid
- Overall dataset is small (n=30) and unbalanced (20 vs. 10) — limits power throughout, especially for Alicante Bouschet-specific claims

:::notes 
Oliver finish 
:::
------------------------------------------------------------------------

## Conclusion: the four questions {.smaller}

Our analysis set out to answer four questions about these 30 wines:

1.  **Can we find groups in the data?**
2.  **Are they consistent with grape type?**
3.  **Can we assume multivariate normality?**
4.  **Are there surprising observations?**

:::notes 

Riley finish 

:::

---

## References {.smaller}
