---
title: "Early Indications of Subtyping within Alicante Bouschet Wines"
subtitle: "AOC Valdeorras, Galicia, Spain — STAT394 Group Report"
author:
  - "Riley Phillips"
  - "Wian de Vos"
  - "Oliver Temel"
  - "Simon Nelson"
format:
  revealjs:
    theme: serif
    slide-number: true
    transition: slide
    width: 1050
    height: 700
    toc: false
    css: |
      .title {
        font-size: 1em !important;
      }
bibliography: references.bib
---

## Background {.smaller}

- Red wine colour, bitterness, astringency based on phenols, including anthocyanins
- Anthocyanins originate in grapes, produced based on genetics
- Ratio of different anthocyanins in wine ("anthocyanin fingerprint") strongly linked to grape cultivar
- Key takeaway: anthocyanin fingerprint of a wine should tell us what grape cultivar it was made  from

---

## The paper {.smaller}

- Compared wine from Mencia grapes (20 samples) with wine from Alicante Bouschet grapes (10 samples)
- UV-Vis Spectrophotometry (9 variables) compared to HPLC-DAD (14 variables)
- Both techniques separated samples by cultivar, HPLC more reliable

---

## The aim {.smaller}

- Look at grouping structure in data from Revilla *et al.*
- Use aggregate data instead of separate blocks

---

## Overview {.smaller}

- Descriptive analysis: cultivars have different means for a number of raw variables
- Correlation analysis: numerous raw variables correlated strongly with each other
- Normality check: several raw variables show evidence of non-normality
- PCA: fixed redundancy, improved normality somewhat
- Mahalanobis distance: 

-> normality check -> PCA -> Mahalanobis distance -> Mahalanobis distance within

---

## The question {.smaller}

- Phenolic composition (anthocyanins, tannins) drives red wine colour, bitterness, astringency, ageing
- Anthocyanin "fingerprint" is under strong genetic control → potentially useful for verifying grape cultivar
- Revilla *et al.* (2016) compared **Mencía** (principal Valdeorras red) vs. **Alicante Bouschet** (teinturier blending grape) — 30 wines, 2014 vintage — using PCA/LDA on spectrophotometric and HPLC blocks **separately**

**Our contribution**: combine all 23 phenolic variables into one joint analysis

- Does the full profile separate cultivars as cleanly as either block alone?
- Is multivariate normality a reasonable assumption?
- Are there atypical wines, or hidden structure within a cultivar?

---

## The data {.smaller}

- 30 wines: 20 Mencía, 10 Alicante Bouschet (2014 vintage, post-malolactic fermentation)
- 23 numeric variables combined from Revilla et al.'s Tables 1, 2, 5, 6:
  - 5 spectrophotometric/colour variables
  - 4 anthocyanin/tannin concentration variables (mg/L)
  - 14 relative-proportion anthocyanins (5 monoglucosides, 9 acylated derivatives)

**Key descriptive pattern**: Alicante Bouschet consistently shows

- ~2× the colour intensity and total phenol index (teinturier character — pigmented flesh, not just skin)
- ~2–3× the anthocyanin/tannin concentrations, but with 2–3× the variability of Mencía


---

## Redundancy among the 23 variables {.smaller}

```{r}
#| echo: false
#| eval: true
#| fig-width: 9
#| fig-height: 8

library(tidyverse)
library(ggcorrplot)

wine <- read_csv("extracted data.csv") |>
  select(-`...1`) |>
  mutate(
    Cultivar = factor(if_else(str_starts(Sample, "M"), "Mencia", "Alicante Bouschet")),
    .after = Sample
  )

wine |>
  select(-Sample, -Cultivar) |>
  cor() |>
  ggcorrplot(
    type = "lower",
    method = "circle",
    lab = TRUE,
    lab_size = 2,
    hc.order = TRUE,
    colors = c("blue", "white", "red"),
    legend.title = "Corr"
  )
```

- **51 of 253 possible pairs (20.2%) correlate at |r| > 0.8** — substantial redundancy
- Near-duplicate measurements: `Tot_Anthocyanins` vs `Tot_Anthocyans`, r = 0.998
- Compositional constraint: monoglucoside proportions sum to ~100%, so they're mechanically correlated (e.g. `MvGl`/`PnGl`, |r| ≈ 0.95–0.97)

> This redundancy motivates dimensionality reduction via PCA


## Is multivariate normality reasonable? {.smaller}

Shapiro-Wilk tested **per variable, per cultivar** (pooling would conflate the cultivar mean gap with real non-normality)

- **12 of 42 within-cultivar tests reject normality** at 5% — far more than the ~2 expected by chance
- Concentrated in:
  - Mencía's total concentration variables (`Tot_Anthocyanins`, `Tot_Anthocyans`, `Tot_Tannins`)
  - Coumaroylated derivatives, in **both** cultivars

> Marginal normality is *necessary* for MVN — these rejections are sufficient to conclude neither cultivar's raw phenolic profile is multivariate normal

**Caveat**: Alicante Bouschet (n=10) has lower power to detect non-normality than Mencía (n=20) — fewer rejections there isn't necessarily "more normal"

---

## Why PCA {.smaller}

Two independent problems point to the same fix:

1. **Singular covariance matrices** — with 23 variables and only 20 (or 10) wines per cultivar, $\hat\Sigma$ has rank ≤ n−1 and can't be inverted → Mahalanobis distance is undefined on the raw variables
2. **Redundancy** — the correlation structure shows much of the 23-dimensional variation is not independent

**Approach**: retain PCs with variance > 1 (Kaiser criterion)

- First 5 PCs retain **90.9%** of total sample variance
- Gives an invertible, lower-dimensional space in which to compute Mahalanobis distances and assess normality

---

## Normality of the principal components {.smaller}

```{r}
#| echo: false
#| eval: false
pander::pander(shapDF)  # PC1–5 Shapiro-Wilk, aggregate + by cultivar
```

- Pooled: **PC1** (p = 0.0001) and **PC5** (p = 0.01) reject normality
- Within cultivar: PC1 and PC5 are fine, **but PC2 rejects in both** cultivars (Mencía p = 0.0161, Alicante Bouschet p = 0.0095)

> Reducing dimensionality helped, but did **not** fully resolve non-normality — Mahalanobis-based p-values later on should be read as approximately, not exactly, valid


---

## Cultivar separation {.smaller}

```{r}
#| echo: false
#| eval: false

wineDataPCADF <- as.data.frame(winePCA$x)
plot(wineDataPCADF$PC1, wineDataPCADF$PC2,
     pch = 21, bg = c(rep("red",20), rep("blue",10)))
```

- PC1–PC2 scatterplot clearly replicates Revilla et al.'s cultivar separation
- Confirmed statistically: **every wine is highly inconsistent with the opposing cultivar's distribution** (p < 0.0001, except one wine at p = 0.0057)
- Within-cultivar Mahalanobis distances flag only a few mildly unusual wines (M7, M8, M17)

> The full 23-variable joint profile separates cultivars just as reliably as Revilla et al.'s separate block-wise analyses

---

## The headline finding: structure within Alicante Bouschet {.smaller}

```{r}
#| echo: false
#| eval: false

wineDataPCADFR = wineDataPCADF[,1:5]
plot(wineDataPCADFR[21:30,]$PC1, wineDataPCADFR[21:30,]$PC2,
     pch = 21, bg = c("green", rep("black",5), rep("green",4)))
```

PC1–PC2 restricted to Alicante Bouschet shows **two distinct clusters**:

- **Cluster 1**: T21, T27–T30 (tight grouping)
- **Cluster 2**: T22–T26 (diverges clearly along PC2)

**Mahalanobis evidence is decisive:**

- Every sample highly consistent with its *own* theorised cluster (p ≥ 0.202)
- Every sample highly inconsistent with the *other* cluster (p < 0.0001)
- Critically, this is **mutual** — not just Cluster 2 wines looking odd from Cluster 1's viewpoint

---

## Why might this be happening? {.smaller}

- Alicante Bouschet already showed 2–3× the variability of Mencía on concentration variables (descriptive stage)
- Revilla et al. attributed this to variable maceration length — the cultivar is often managed to different blending targets
- **Our finding refines this**: rather than one heterogeneous group with a long tail, the data are more consistent with **two discrete phenolic profiles**

> Plausible explanation: different maceration regimes or blending targets across batches — but we have no winemaking records to confirm this directly

---

## Limitations {.smaller}

- Subgroup analysis rests on **5 wines per cluster**, tested against covariance estimated from those same 5 observations (3 PCs) — small enough to be suggestive, not confirmatory
- PC2's non-normality means the chi-square reference distribution underlying all Mahalanobis p-values is only **approximately** valid
- Overall dataset is small (n=30) and unbalanced (20 vs. 10) — limits power throughout, especially for Alicante Bouschet-specific claims

---

## Conclusion {.smaller}

- The combined 23-variable phenolic profile reliably separates Mencía from Alicante Bouschet, replicating Revilla et al. without needing separate variable blocks
- **Alicante Bouschet itself appears phenolically non-homogeneous** — splitting into two sub-clusters with a separation comparable to the between-cultivar separation
- Given small sample size, this is a **hypothesis-generating result**, not a confirmed finding — worth testing in a larger sample with known maceration/blending records

Questions?

